---
title: "Tentadokument"
author: "Gustav Sternelöv"
date: "18 oktober 2016"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    highlight: tango
---

\pagebreak

# Lab 1
## Group report

### Assignment 1
```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(bnlearn)
# Assignment 1
data("learning.test")
data("alarm")
data("asia")

set.seed(1897)

init_graph <- random.graph(LETTERS[1:6])
res1 <- hc(learning.test, start = init_graph, restart = 100)
```
We begin by using the *Hill-climbing algorithm* to setup the strucutre of the network.
We then use this structure in the *Hill-climbing algorithm* and set the number of restarts equal to 100.


```{r, echo=TRUE}
all.equal(cpdag(init_graph), cpdag(res1))
```

The function *all.equal* can be used to compare the different networks, we here get the result that they are not the same. This can also be seen in the two graphs below.

```{r, echo=TRUE, fig.width=12}
plot(init_graph, main="DAG 1")
```

```{r, echo=TRUE, fig.width=12}
plot(res1, main="DAG 2, with 100 restarts")
```

We notice that we receive two different structures which represent two different models.

```{r, echo=TRUE}
score(init_graph, learning.test)
score(res1, learning.test)
```
The score of the two dags is seen above. We notice that they recive different scores which is logical since the DAGs are different. However, the score is not a sufficient value to use for comparing if two DAGs are equal since it is possible for two different DAGs to obtain the same score. This is a consequence of the independencies, DAGs with the same independencices will always be equivalent if the score is used for comparing them.

### Assignment 2
```{r, echo=TRUE}
# Assignment 2
library(ggplot2)
res_asia <- hc(asia)
res1_asia <- hc(asia, start = res_asia, restart = 100)
```

We examine the effect of decreasing the regularization by looking at the complexity of the model given different imaginary sample size. Below are the plots for ISS equal 10, 100 and 1000. It can be seen that as the ISS increases the effect is that more complex models are recieved.  

```{r, echo=TRUE, fig.width=12}
p1 <- hc(asia, start=res_asia, restart = 100, score="bde", iss=10)
p2 <- hc(asia, start=res_asia, restart = 100, score="bde", iss=100)
p3 <- hc(asia, start=res_asia, restart = 100, score="bde", iss=1000)

par(mfrow=c(1,3))
plot(p1, main="ISS: 10")
plot(p2, main="ISS: 100")
plot(p3, main="ISS: 1000")
par(mfrow=c(1,1))
```
When the number of ISS increases we obtain more dense DAGs. However these changes becomes less evident when the ISS becomes large. This is due to that the regularization has decreased. The explanation of this occurence lies in how the parameter for the penalty has changed. Since the model is less regularized the penalty must have been lowered, and as a result the models become more complex. This is not wanted as a complex model may be heavily overfitted to data and therefore the model normally is regularized more than in the examples here. 


### Assignment 3
```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(gRain)
# Assignment 3
#true data
Init2 <- hc(asia,restart = 100)
fit <- bn.fit(x = Init2,data =asia)
#exact value
grainOBJ<-as.grain(fit)
compGrainOBj<-compile(grainOBJ)
compGrainOBj2<-setFinding(compGrainOBj,nodes = c("A"),states = c("no"))
compGrainOBj4<-setFinding(compGrainOBj,nodes = c("A","B","E"),states = c("no","yes","no"))
compGrainOBj6<-setFinding(compGrainOBj,nodes = c("A","B","E","D","T"),states = c("yes","yes","no","no","no"))
prob1 <- querygrain(compGrainOBj2,type = "marginal") 
prob2 <- querygrain(compGrainOBj4,type = "marginal")
prob3 <- querygrain(compGrainOBj6,type = "marginal")
exaxt1 <- c(prob1$B[2],prob2$D[2],prob3$S[2])


#approximate data
#approximate answer will be bad when we have many nodes since it will give us very few observation
set.seed(11827)
compAprox2<-cpdist(fit,nodes =c("A","B"),evidence = (A=="no"))
set.seed(11827)
compAprox4<-cpdist(fit,nodes =c("A","B","E","D"),evidence = (A=="no"&B=="yes"&E=="no"))
set.seed(11827)
compAprox6<-cpdist(fit,nodes =c("A","B","E","D","T","S"),evidence = (A=="yes"&B=="yes"&E=="no"&D=="no"& T=="no"))

approx <- c(prop.table(table(compAprox2$B))[2], prop.table(table(compAprox4$D))[2], prop.table(table(compAprox6$S))[2])          

data.frame("Obs.Nodes" = c("A", "A,B,E", "A,B,E,D,T"),
           "Statement " = c("No", "No,Yes,No", "y,y,n,n,n"),
           "Next included" = c("B","D","S"), "Statement"=c(
             "Yes","Yes","Yes"), "True prob"=round(exaxt1,5), 
           "Approx prob"=round(approx,5), "Diff"=round(abs(exaxt1-approx), 5))

```

One can see that the approximated algorithm get less efficient when number of observed nodes increases, this can be seen in the "diff" column where the difference increases. This is an expected outcome, since more observed nodes leads to that there is lesser observations that satisfies the decided statement. So theoretically given zero observated nodes the approximated algorithm and LS should be equally good.

In the table below are the same query for the approximate method performed three times.

```{r, echo=TRUE}
compAprox2<-cpdist(fit,nodes =c("A","B"),evidence = (A=="no"))
prop.table(table(compAprox2))
           
compAprox2<-cpdist(fit,nodes =c("A","B"),evidence = (A=="no"))
prop.table(table(compAprox2))
           
compAprox2<-cpdist(fit,nodes =c("A","B"),evidence = (A=="no"))
prop.table(table(compAprox2))
```

The function *cpdist* are based on Monte Carlo particle filters, which uses a random simulation and therefore it is possible to reccieve different result for each run.

### Assignment 4
```{r, echo=TRUE}
# Assignment 4

fraction <- data.frame(Every1=NA, Every10=NA,Every20=NA)

Every <- c(1,10,20)
burn <- c(300, 1000, 5000)
countCol <- 0

for(i in Every){
  countCol <- countCol + 1
  countRow <- 0
  for(j in burn) {
    countRow <- countRow + 1
    
    DAGSGraph<-random.graph(LETTERS[1:5],method = "melancon",num=30000,every=i,burn.in=j)
cpFraction<-list()
for(k in 1:length(DAGSGraph)){
  cpFraction[[k]]<-cpdag(DAGSGraph[[k]])  
}
  fraction[countRow, countCol] <- length(unique(cpFraction))  / 30000  
  }
}
row.names(fraction) <- c("Burn300", "Burn1000", "Burn5000")
```

Depending on the chosen settings for the *random.graph* function a different number of unique EGs are returned. In the table below are the results for different values of every and burn-in summarized. The burn-in parameter does not seem to affect the fraction of EGs returned. Every on the other hand do have some effect since a higher value increase the number of EGs. Especially, for a value higher than one larger fractions are returned. 

```{r, echo=TRUE}
fraction
```

Since around 25 % of the DAGs represent different independence models the structure learning becomes a bit problematic. This is because the same essential graph in many cases then represent several different DAGs and it is not possible to determine which of these DAGs that is the most appropriate one. The effect of this is that it is hard to find the true structure for a network, more precise the true casualities in the actual network. 



## Individual report

```{r, fig.width=12}
set.seed(311015)
TestData <- data.frame(ground=rep(c("H","A"),500,each=10),
                       surface=rep(c("Artif", "Grass"),1000, each=5),
  goals=sample(c(0,2,3,4,5), 10000, replace=TRUE,prob=c(2/9,2/6,1/6,1/6,1/6)),
  bookings=sample(c(0,1,2,3), 10000, TRUE, rep(1/4,4)))

TestData$Res = 0
TestData$Red = 0           
set.seed(311015)
for (i in 1:nrow(TestData)){
  if(TestData$ground[i] == "H" & TestData$surface[i] == "Artif"){
    TestData$Res[i] = sample(c("1","X","2"),1,prob=c(2/3,1/6,1/6))    
  }
  if(TestData$ground[i] == "A" & TestData$surface[i] == "Artif"){
    TestData$Res[i] = sample(c("1","X","2"),1,prob=c(1/6,1/6,2/3)) 
  }
  if(TestData$ground[i] == "H" & TestData$surface[i] == "Grass"){
    TestData$Res[i] = sample(c("1","X","2"),1,prob=c(3/6,2/6,1/6)) 
  }
  if(TestData$ground[i] == "A" & TestData$surface[i] == "Grass"){
    TestData$Res[i] = sample(c("1","X","2"),1,prob=c(1/6,3/6,2/6)) 
  }
  if(TestData$goals[i] == 0){
    TestData$Res[i] = "X"
  }else{
    TestData$Res[i] = TestData$Res[i] 
  }
  if(TestData$bookings[i] > 2){
    TestData$Red[i] = 1
  }else{
    TestData$Red[i] = 0
  }
}

TestData$Res <- as.factor(TestData$Res)
TestData$ground <- seq_along(levels(TestData$ground))[TestData$ground]
TestData$surface <- seq_along(levels(TestData$surface))[TestData$surface]
TestData$Res <- seq_along(levels(TestData$Res))[TestData$Res]
TestData$ground <- as.factor(TestData$ground)
TestData$surface <- as.factor(TestData$surface)
TestData$Res <- as.factor(TestData$Res)
TestData$goals <- as.factor(TestData$goals)
TestData$bookings <- as.factor(TestData$bookings)
TestData$Red <- as.factor(TestData$Red)
# Lab 1

## Assignment 1
# Initialize a random graph with 5 nodes
set.seed(311015)
init_graph <- random.graph(colnames(TestData))
# Run the score based structure learning algorithm with 100 restarts
res1 <- hc(TestData, start = init_graph, restart = 100)
all.equal(cpdag(init_graph), cpdag(res1))
plot(init_graph, main = "DAG 1")
plot(res1, main = "DAG 2, with 100 restarts")
score(init_graph, TestData)
score(res1, TestData)

## Assignment 2
# Run the score based learning algorithm for 3 different imaginary sample sizes
# initial graph is the graph returned in assignment 1

p1 <- hc(TestData, start = res1, restart = 100, score = "bde", iss = 10)
p2 <- hc(TestData, start = res1, restart = 100, score = "bde", iss = 100)
p3 <- hc(TestData, start = res1, restart = 100, score = "bde", iss = 1000)
par(mfrow = c(1, 3))
plot(p1, main = "ISS: 10")
plot(p2, main = "ISS: 100")
plot(p3, main = "ISS: 1000")
par(mfrow = c(1, 1))

# In general, number of connections increases with iss. 
# Results in a model that finds non-existing connections. 
# Overfitted to this data, this sample. 


## Assignment 3
# creates object for approximate method from res1
res1Fit <- bn.fit(res1, TestData)

# approximate method, no evidence
evi_cb <- cpdist(res1Fit, nodes = c("surface"), evidence = TRUE)
prop.table(table(evi_cb))

# Exact method, no evidence
# object for exact method
res1Fit_grain <- as.grain(res1Fit)

res1Fit_grain <- compile(res1Fit_grain)
res1grain <- setFinding(res1Fit_grain,nodes=c("surface"),
                          states = TRUE)
# Which node that is shown below is determined by [i]
querygrain(res1grain)[2]

# Clearly see difference between methods when looking at ground or surface
# for which the proportions are known. 

# Approximate method, with evidence
evi_cb <- cpdist(res1Fit, nodes = c("goals"), evidence = (surface == "2"))
prop.table(table(evi_cb))

# Exact method, with evidence
ltFit_grain <- setEvidence(res1Fit_grain,nodes=c("surface"), states = c("2") )
querygrain(ltFit_grain)[3]

# Still similar results from the respective methods #

# Approxiamte method for several different seeds
set.seed(0814)
evi_cb <- cpdist(res1Fit, nodes = c("Red"), evidence = TRUE)
prop.table(table(evi_cb))

set.seed(311015)
evi_cb <- cpdist(res1Fit, nodes = c("Red"), evidence = TRUE)
prop.table(table(evi_cb))

set.seed(1991)
evi_cb <- cpdist(res1Fit, nodes = c("Red"), evidence = TRUE)
prop.table(table(evi_cb))

set.seed(1897)
evi_cb <- cpdist(res1Fit, nodes = c("Red"), evidence = TRUE)
prop.table(table(evi_cb))

# As expected, differet results are obtained for different seeds

# Approximate versus Exact when many observed nodes
evi_cb <- cpdist(res1Fit, nodes = c("Res"),evidence=c(ground=="2" & surface=="1" &
                                           goals=="3" & bookings=="2" & Red=="0"))
prop.table(table(evi_cb))

# Exact method, no evidence
ltFit_grain <- setFinding(res1Fit_grain,nodes=c("ground","surface","goals", "bookings", "red"),states = c("2","1", "3", "2", "0"))
querygrain(ltFit_grain)[4]

# Results differs more when more observed nodes
# Problem for approximate method: Fewer and fewer observations that matches the evidence
# when the number of observed nodes increases
```




\pagebreak


# Lab 2

## Group report

```{r, echo=TRUE, warning=FALSE, message=FALSE}
library(HMM)
library(ggplot2)
library(scales)
library(gridExtra)
```

### 1

```{r, echo=TRUE}
library(ggplot2)
# Assignment 1
# SET UPP MY HMM OBJECT
library(HMM)

# Set up my transmission matrix
# P of leaving and P of staying in a given state is 0.5.
# Robot can only move forward.

# The bad GPS gives a prob of 0.2 to be between i-2:i+2.
trans<-matrix(rep(0,100),nrow = 10)
emission<-matrix(rep(0,100),nrow = 10)
for(i in 1:ncol(trans)){
  trans[i,i]<-0.5
  if((i+1)%%10==0){
    trans[i,i+1]<-0.5
  }
  trans[i,(i+1)%%10]<-0.5
}
for(i in 1:ncol(emission)){
  emission[i,c(i-2,i-1,i,i+1,i+2)%%10]<-0.2
  if(sum(emission[i,])!=1){
    emission[i,ncol(emission)]<-0.2
  }
}
# Set up my Hiden markov model
# Symbols=hidden layers?
set.seed(11827)
hmm <- initHMM(States= 1:10, Symbols = 1:10,
               startProbs = rep(0.1,10), transProbs = trans,
               emissionProbs = emission)
```


According to the scenario given in the lab description we build the model by this code:

```{r, echo=TRUE}
# The start is randomly choosen with 10% for each state.
hmm <- initHMM(States= 1:10, Symbols = 1:10,
               startProbs = rep(0.1,10), transProbs = trans,
               emissionProbs = emission)
```

The robot has a chance of staying of 50 % and 50 % of moving one state up.
The GPS of the robot is of poor quailty therefore the transmission matrix will have 20 % probability for all states +- 2 states from the current state.
We set the starting probability equal for all the states.

### 2 
```{r, echo=TRUE, fig.height=3.5, fig.width=7}
###### Assignment 2
# Simulate given hmm and n.
  set.seed(311015)
  simulateHMM <- simHMM(hmm, 100) # n is sample size
  
  sim_2 <-  data.frame(Time=1:100, Path=simulateHMM$states) # DF for ggplot
  
  # Plotting simulated states with time.
  ggplot(sim_2) + geom_line(aes(x=Time, Path))  +
  scale_x_continuous(breaks=seq(0,100,10)) +
  scale_y_continuous(breaks=1:10) +
    theme(panel.grid.minor = element_blank(), axis.title.y = element_text(angle=0)) +
    labs(y="States", x="Steps", title="Hidden states")
  
```

We simulate a HMM for 100 time steps.For this specific simulation we started at state six, and ended at state seven after 100 steps.
As expected,the robot does not take any steps backwards, either it stays or it moves to a higher state.

### 3-4
```{r, echo=TRUE}
###### Assignment 3-4
# Function to calculate probabale path given probabilities matrix(states x time).
countProb <- function(x, states, n) {
  mostProb <- c()
  for(i in 1:n) {
    mostProb[i] <- which.max(x[states,i] )
  }
  return(mostProb)
}
# START FUNCTION
# Require a hmm object.
# This function returns accuarcy of smoothing and filtering method.
# hmm is initHMM object, n =samplem and states=Vector of states.
MYFUNC <- function(hmm, n=100, states) {
  
  # options(scipen = 999) # Used to keep numbers as numbers.
  ############################################################  
  # This is assignment 2  
  ############################################################  
  # Simulate given hmm and n.
  set.seed(311015)
  simulateHMM <- simHMM(hmm, n) # n is sample size
  
  ############################################################  
  # This is assignment 3  
  ############################################################
  # Calculate Alpha and Beta matrix. Use exp to get real prob insted of logged.
  set.seed(173113)
  alpha <- prop.table(exp(forward(hmm, simulateHMM$observation)))
  
  for(i in 1:n) {
  alpha[, i] <- alpha[,i]/colSums(alpha)[i]
  }
  
  # Calculate smooth probabilities.
  smooth<- posterior(hmm, simulateHMM$observation) # Smooth
  
  # Most probable path using viterbi algorithm
  probPathViterbi <- viterbi(hmm, simulateHMM$observation)
  ############################################################  
  # This is assignment 4  
  ############################################################
  # Set all information in big data.frame
  bigDF <- as.data.frame(cbind(probPathSmoothing = countProb(smooth, states = states, n=n),
                               probPathFiltering = countProb(alpha, states = states, n=n),
                               probPathViterbi = probPathViterbi,
                               Simulated = simulateHMM$states,
                               accSmooth = c(0),
                               accFiltering = c(0),
                               accViterbi= c(0)))
  
  # Set all equal simulated path to 1. Then I can use mean() to calculate Accuracy.
  bigDF[bigDF$probPathSmoothing==bigDF$Simulated, "accSmooth"] <- 1
  bigDF[bigDF$probPathFiltering==bigDF$Simulated, "accFiltering"] <- 1
  bigDF[bigDF$probPathViterbi==bigDF$Simulated, "accViterbi"] <- 1
  
  # Return a list consisting of three accuarcy, paths and a simulated vector.
  resList <- list(AccuarySmooth= mean(bigDF$accSmooth),
       AccuarcyFiltering=  mean(bigDF$accFiltering),
       AccuarcyViterbi=  mean(bigDF$accViterbi),
       probPathSmoothing= bigDF$probPathSmoothing,
       probPathFiltering= bigDF$probPathFiltering,
       probPathViterbi= bigDF$probPathViterbi,
       Simulates= bigDF$Simulated,
       alpha=alpha)
  
  return(resList)
}

```

To solve this assignment we choose to  make a function to be able to calculate the filtering algorithm. This is done by using the following formula:

$$Filtering: \: \frac{\alpha(z^t) }{\sum_{z^t} \alpha (z^t)}$$

For the smoothing probabilities we use the function *posterior()*. Which calculates accordingly to this following formula:

$$Smoothing: \: \frac{\alpha(z^t) \beta(z^t)}{\sum_{z^t} \alpha (z^t) \beta(z^t)}$$

And finally we compute the most probable path using the *Viterbi* algorithm. 

The accuarcy is shown below.

```{r, echo=FALSE}

# Print accuarcy
unlist(MYFUNC(hmm = hmm, n = 100, states=1:10)[1:3])

```

Here we notice that the accuarcy is highest for the smoothed distribution and is the lowest for the most probable path.

We plot the predicted paths and investigate them further.

```{r, echo=FALSE, fig.height=3.5, fig.width=7}
# Plotting the paths to compare
test <- MYFUNC(hmm = hmm, n = 100, states=1:10)
path_df <- data.frame(Path=c(test$probPathSmoothing, test$probPathFiltering, test$probPathViterbi),
                      Method= rep(c("Smoothing", "Filtering", "MPP"),each=100),
                      x = rep(1:100,3))

# Plot the different paths
ggplot() + geom_line(data=path_df, aes(x=x, y=Path, color=Method), size=1) +
  scale_x_continuous(breaks=seq(0,100,10)) +
  scale_y_continuous(breaks=1:10) +
  theme(panel.grid.minor = element_blank(), axis.title.y = element_text(angle=0)) +
  labs(title="Probable paths for Smoothing, Filtering and most probable path(MPP)",
       y="States",
       x="Steps")
```

We notice that the different paths start and end in different states.
One intresting finding is that the Filtered path at least moves backwards four times, this is not allowed and therefore we suspect that this distribution will not give the most optimal result.
It can also be seen that the Viterbi algorithm returns a path that get stuck for long sequences, especially for state one.

### 5
```{r, echo=TRUE, fig.height=3.5, fig.width=7}
###### Assignment 5
# Investigate how accuarcy changes for larger sample sizes
n <- c(5,20,50,80,100,120,150,175,200,225,250,275,300) # These are the sample sizes 
# DF for plotting
df <- data.frame(Value=c(NA),
                 Method=rep(c("MPP", "Filtering" ,"Smoothing"), each=length(n)),
                 x=rep(n, 3))

# These values are needed as a consequence of ggplot plotting
# k, j, l is the 
k <- 0
j <- length(n)
l <- length(n) * 2
for(i in n) {
  k <- k+1
  j <- j+1
  l <- l+1
  temp <- MYFUNC(hmm = hmm, n = i, states=1:10)
  df[k, "Value"] <-  temp[3] # Smooth acc
  df[j, "Value"] <-  temp[2] # Filtering acc
  df[l, "Value"] <-  temp[1] # MPP acc
}

ggplot() +
  geom_line(data=df , aes(x=x, y=Value , color=Method )) +
  ylim(c(0,1))

```

The three diffeerent accuracies is visualized in the above plot. This is done for different sample sizes. We noticed that the smoothed accuracies is superior to the two others for all sample sizes. At at a glance of the plot one can say that both the filtering and MPP obtain similiar accuarcy.

The explanation of the smoothed superior behavior compared to the filterd accuarcy is that that the smoothed uses more data. This can be seen in that formulas in assignment 3, where the smoothing algorithm uses both the backward and forward part of the algorithm, compared to the filtered algorithm which only uses the forward part.

The smoothing distribution is obtained by the *Forward-Backwards algorithm* while the most probable path is obtained by the *Viterbi algorithm*. The Forward-backwards algorithm gives marginal probability for each individual state while the Viterbi gives probability for the most likely sequence of states.

### 6
```{r, echo=TRUE, fig.height=3.5, fig.width=7}
###### Assignment 6
library(entropy)
current.model <- MYFUNC(hmm = hmm, n = 100, states=1:10)
Estimated_Entropy <- c()

for(i in 1:100) {
  Estimated_Entropy[i] <- entropy.empirical(current.model$alpha[,i])
}
gg_entropy <- data.frame(Entropy= Estimated_Entropy,
                         Time.Steps=1:100)
ggplot() + geom_line(data=gg_entropy, aes(x=Time.Steps, y=Entropy))

```

Based on the apperance of the graph we can see that increasing time steps does not affect the trend or evidence of the robots position.
For low values of the empirical entropy we are more sure about the position of the robot.


### 7
```{r, echo=TRUE}
###### Assignment 7
set.seed(2131)
sim <- simHMM(hmm, 100)
alpha <- prop.table(exp(forward(hmm, sim$observation)))
filtering <- alpha/colSums(alpha)
filtering[,100] %*% trans / sum(filtering[,100] %*% trans) # We know state in t100=3
```

The most probable state for t=101 for this simulation is that the robot will be in state 5.
The vector above is calculated by the following formulas.

$$p(Z^{101}|X^{0:100}) = \sum\limits_{Z^{100}} p(Z^{101}, Z^{100} | X^{0:100})$$
$$p(Z^{101}|Z^{100}, X^{0:100}) p(Z^{100}|X^{0:100}) = p(Z^{101}|Z^{100}) p(Z^{100}|X^{0:100})$$

Basically this is just the transition matrix times our previous estimated filtering distribution.

\pagebreak



## Individual report
```{r, fig.height=3.5, fig.width=7}

# 1. Initialize the model
# Transition matrix is unchanged
# Emission matrix is modified, see below

seq_eP <- rep(1:10,10)
transP <- matrix(data=0, nrow=10, ncol=10)
for(i in 1:10){
  transP[i, c(seq_eP[i], seq_eP[i+1])] <- 0.5
}
# Emission matrix only give probabilities for states +-1
emissionP <- matrix(data=0, nrow=10, ncol=10)
for(i in 1:10){
  j <- i+10
  emissionP[c(seq_eP[j-1],seq_eP[j], seq_eP[j+1]),i] <- 0.2
}

HMM1 <- initHMM(States = c(1:10), Symbols = c(1:10), startProbs =  rep(0.1, 10),
                transProbs = transP, emissionProbs = emissionP)

# 100 observations are simulated from the model
set.seed(311015)
Sim1 <- simHMM(HMM1, 100)

# A plot over the simulated states and the observations
Sims1 <- data.frame(Time=rep(1:100, 2), Data=c(Sim1$states, Sim1$observation),
                    Var=rep(c("State", "Observation"), each=100))
ggplot(Sims1, aes(x=Time, y=Data, col=Var)) + geom_line(size=1) + theme_classic()
# State sequence never moves backwards but observation sometimes does


# The filtered distribution is computed
options(scipen=99)
alpha <- exp(forward(HMM1, Sim1$observation))
alpha <- prop.table(alpha)
filtFunc <- function(Data){
  Data <- Data / sum(Data)
}
filtering <- apply(alpha,2,filtFunc)
probState <- function(Data){
  Data <- which.max(Data)
}
Filter <- data.frame(state=apply(filtering, 2, probState), Time=1:100)

filter_p <- ggplot(Filter, aes(x=Time, y=state)) + geom_path() +
  labs(x="Path", y="State",title="Filtered distribution") +
  scale_y_continuous(breaks=c(1:10)) + theme_classic()

# Smoothed distribution
smoothing <- posterior(HMM1, Sim1[[2]])
smooth <- data.frame(state=apply(smoothing, 2, probState), Time=1:100) 
smooth_p <- ggplot(smooth, aes(x=Time, y=state)) + geom_path() +
  labs(x="Path", y="", title= "Smoothed distribution") +
  scale_y_continuous(breaks=c(1:10)) + theme_classic()

# The most probable path
path <- data.frame(state=as.numeric(viterbi(HMM1, Sim1[[2]])), Time=1:100)
path_p <- ggplot(path, aes(x=Time, y=state)) + geom_path() +
  labs(x="Path", y="", title="Most probable path") +
  scale_y_continuous(breaks=c(1:10)) + theme_classic()

# All distributions next to each other
plots <- list(filter_p, smooth_p, path_p)
plots1 <- arrangeGrob(grobs = plots, nrow=1)
plot(plots1)

# All distributions in the same plot
Filter$distribution <- "Filtered"
smooth$distribution <- "Smoothed"
path$distribution <- "Path"
allDist <- rbind(Filter, smooth, path)
ggplot(allDist, aes(x=Time, y=state, col=distribution)) + geom_path(size=1) +
  labs(x="Path", y="State", title="Comparsion of the distributions") +
  scale_y_continuous(breaks=c(1:10)) + theme_classic()

# The accuracy for the respective distribution
data.frame(AccuracyFiltered = sum(Sim1[[1]] == Filter$state) / 100,
           AccuracySmoothed=sum(Sim1[[1]] == smooth$state) / 100,
           AccuracyPath= sum(Sim1[[1]] == path$state) / 100)

# The frequency for each state for all distributions
# Not sure if this says something of importance
data.frame(Filtered=prop.table(table(Filter$state))[1:10],
           Smoothed=prop.table(table(smooth$state))[1:10],
           Path=prop.table(table(path$state))[1:10])


# The accuracy with different samples
# Samples 100 observations 1000 times and calculate the accuarcy for each sample
accuracy <- data.frame(sampleS=1:400, accF=0, accS=0, accP=0)
set.seed(0814)
for(t in 1:nrow(accuracy)){
  Sim2 <- simHMM(HMM1, 100)
  #Filtering
  alpha <- exp(forward(HMM1, Sim2$observation))
  alpha <- prop.table(alpha)
  filtering <- apply(alpha,2,filtFunc)
  Filter <- data.frame(state=apply(filtering, 2, probState), Time=1:100)
  # Smoothing
  smoothing <- posterior(HMM1, Sim2[[2]])
  smooth <- data.frame(state=apply(smoothing, 2, probState), Time=1:100) 
  # MPP
  path <- data.frame(state=as.numeric(viterbi(HMM1, Sim2[[2]])), Time=1:100)
  accuracy[t,2] <- sum(Sim2[[1]] == Filter$state) / 100
  accuracy[t,3] <- sum(Sim2[[1]] == smooth$state) /100
  accuracy[t,4] <- sum(Sim2[[1]] == path$state) / 100
}
# tidyr to transform data to ggplot format
accuracy2 <- tidyr::gather(accuracy, "method", "accuracy", 2:4)
ggplot(accuracy2, aes(x=sampleS, y=accuracy, col=method)) + geom_line( size=1)+
  theme_classic()+scale_y_continuous(labels=percent) +
  scale_x_continuous(breaks=seq(50,350,50))
# Smoothed the most accurate. Very even between filtered and MPP

# The entropy for each time step for the filtered distribution
set.seed(1897)
Sim3 <- simHMM(HMM1, 100)
alpha <- prop.table(exp(forward(HMM1, Sim3$observation)))
filtering <- apply(alpha,2,filtFunc)
Estimated_Entropy <- c()
for (i in 1:100) {
  Estimated_Entropy[i] <- entropy.empirical(filtering[, i])
}
gg_entropy <- data.frame(Entropy = Estimated_Entropy, Time = 1:100)
ggplot(gg_entropy, aes(x = Time, y = Entropy)) + geom_line()

# Higher value indicates higher uncertainty over where the robot is.
# More sure for lower values and knows exactly where it is if entropy = 0
# Can see that the certainty for the robot's position not increases with time.
# That we knew the position in the previous time step does not necessarily helps us
# in the next time step. 


# For the simulation in the previous assignment are the probabilities for state 101
# calculated. This is done by using the probabilities from the filtering for the 
# 100th simulation and the whole transition matrix

filtering[, 100] %*% transP/sum(filtering[, 100] %*% transP) 

# Highest probability for Z101 = 6. That is the individually most probable state
# for time step 101. 

```



\pagebreak


# Lab 3

## Group report

### Assignment 1
#### a)
The implementation of a function that simulates from the posterior distribution *f(x)* by using the squared exponential kernel is done in two steps. In the first step a function that computes the squared exponential kernel is created. 
The formula for the squared exponential kernel can be seen below:  
$$K(x, x') = \sigma_f^2 \times exp(-0.5 \times(\frac{x-x'}{\iota})^2)$$

The second step is to build the function *PosteriorGP*. The aim with this function is to calculate the posterior mean and variance of *f* over a grid of x-values. This is done by implementing algorithm 2.1 from the book *Gaussian Processes for Machine Learning* (Rasmussen and Williams). 

The code that has been used to implement the functions can be seen in the appendix *R-code*.  

The prior mean of *f* is assumed to be zero for all *x*, which gives the following prior distribution for *f(x)*:
$$f(x) \sim GP(0, K(x, x'))$$  

Then, the posterior guassian distribution looks as following:  
$$f_* \mid x,y,x_* \sim N(\bar{f_*},cov(\bar{f_*})$$  

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(ggplot2)
# Lab 3
# Assignment 1
SqExpKernel <- function(x1, x2, hyperParam){
  K <- matrix(nrow=length(x1), ncol=length(x2))
  for (i in 1:length(x2)){
    K[, i] <- hyperParam[1]^2 * exp(-0.5 *( (x1-x2[i])/hyperParam[2]) ^2)
  }
  return(K)
}
PosteriorGP <- function(x, y, xStar, hyperParam, sigmaNoise){
  # Calculates f star bar
  K <- SqExpKernel(x, x, hyperParam) 
  L <- t(chol(K + sigmaNoise*diag(dim(K)[1]) )) 
  alpha <- solve(t(L),solve(L,y))
  # Posterior mean
  fStar <- (SqExpKernel(xStar, x, hyperParam)) %*% alpha 
  # Posterior variance
  v_f <- solve(L, t(SqExpKernel(xStar, x, hyperParam)))
  cov_fStar <- SqExpKernel(xStar, xStar, hyperParam) - (t(v_f) %*% v_f)  
  # Store all values in a list
  val_list <- list(fStar=fStar, xStar=xStar, cov_fStar=cov_fStar, xStar=xStar
  )
  return(val_list)
}
```


#### b)
Since the noise standard deviation is assumed to be known the parameter $\sigma_n$ is set to 0.1. The prior hyperparameter $\sigma_f$ is set to 1 and the second prior hyperparameter $\iota$ is set to 0.3. Furthermore the prior is updated with one observation, *(x,y)*=(0.4, 0.719). A plot over the posterior mean of *f* over the interval *x* $\in$ [-1,1] with 95 % probability bands for *f* can be seen below.  

```{r, echo=TRUE, fig.height=2.75, fig.width=7, fig.align='center'}
assign1B <- PosteriorGP(x=0.4, y=0.719, xStar=seq(-1,1, 0.01), hyperParam=c(1, 0.3),
            sigmaNoise=0.1^2)
Upper1B <- assign1B$fStar + 1.96 * sqrt(diag(assign1B$cov_fStar))
Lower1B <- assign1B$fStar - 1.96 * sqrt(diag(assign1B$cov_fStar))
plotD <- data.frame(fStar=assign1B$fStar, xStar=assign1B$xStar, Lower=Lower1B, Upper=Upper1B)
xy <- data.frame(x=0.4, y=0.719)
ggplot(plotD, aes(y=fStar, x=xStar)) + geom_point(col="darkorange") + ylim(-2,2.2) + 
  geom_line(data=plotD, aes(xStar, Lower), col="seagreen", size=1.1) +
  geom_line(data=plotD, aes(xStar, Upper), col="seagreen", size=1.1) +geom_point(data=xy, aes(x,y), size=3) + theme_classic()
```
  
  The black dot is the observed value and it can be seen that the probability bands are more narrow around this value.

#### c)
The posterior from *b)* is updated with another observation, *(x,y)*=(-0.6, -0.044).  
```{r, echo=TRUE, fig.height=2.75, fig.width=7}
assign1C <- PosteriorGP(x=c(0.4, -0.6), y=c(0.719, -0.044), xStar=seq(-1,1, 0.01), hyperParam=c(1, 0.3),sigmaNoise=0.1^2)
Upper1C <- assign1C$fStar + 1.96 * sqrt(diag(assign1C$cov_fStar))
Lower1C <- assign1C$fStar - 1.96 * sqrt(diag(assign1C$cov_fStar))

plotD <- data.frame(fStar=assign1C$fStar, xStar=assign1C$xStar, Lower=Lower1C, Upper=Upper1C)
xy <- data.frame(x=c(0.4, -0.6), y=c(0.719, -0.044))
ggplot(plotD, aes(y=fStar, x=xStar)) + geom_point(col="darkorange") + ylim(-2,2.2) + 
  geom_line(data=plotD, aes(xStar, Lower), col="seagreen", size=1.1) +
  geom_line(data=plotD, aes(xStar, Upper), col="seagreen", size=1.1) +geom_point(data=xy, aes(x,y), size=3) +theme_classic()
```  
  
  Again it can be seen that the probability bands are more narrow around the observed values, and that they are quite wide for the other values. 

#### d)
In *d)* the number of observations rises to five, resulting in the following plot over the posterior mean of *f* and its 95 % probability intervals.  
```{r, echo=TRUE, fig.height=2.75, fig.width=7}
assign1D <- PosteriorGP(x=c(0.8, 0.4, -0.2, -0.6, -1), y=c(-0.664, 0.719, -0.94, -0.044, 0.768), xStar=seq(-1,1, 0.01),hyperParam=c(1, 0.3),sigmaNoise=0.1^2)
Upper1D <- assign1D$fStar + 1.96 * sqrt(diag(assign1D$cov_fStar))
Lower1D <- assign1D$fStar - 1.96 * sqrt(diag(assign1D$cov_fStar))

plotD <- data.frame(fStar=assign1D$fStar, xStar=assign1D$xStar, Lower=Lower1D, Upper=Upper1D)
xy <- data.frame(x=c(0.8, 0.4, -0.2, -0.6, -1), y=c(-0.664, 0.719, -0.94, -0.044, 0.768))
ggplot(plotD, aes(y=fStar, x=xStar)) + geom_point(col="darkorange") + ylim(-2,2.2) + 
  geom_line(data=plotD, aes(xStar, Lower), col="seagreen", size=1.1) +
  geom_line(data=plotD, aes(xStar, Upper), col="seagreen", size=1.1) +geom_point(data=xy, aes(x,y), size=3) +theme_classic()
```
  
Compared to the plots in *b)* and *c)*, the curve for the posterior mean of *f* is less straight/ more curvaceous than before. The probability bands has also changed and are thanks to the rise from two to five observed values more narrow, but also quite wiggly. 

#### e)
The hyperparameter $\iota$ is now set to 1. The other parameters are unchanged and the same observations as in *d)* are used.  
```{r, echo=TRUE, fig.height=2.75, fig.width=7}
assign1E <- PosteriorGP(x=c(0.8, 0.4, -0.2, -0.6, -1), y=c(-0.664, 0.719, -0.94, -0.044, 0.768), xStar=seq(-1,1, 0.01),hyperParam=c(1, 1),sigmaNoise=0.1^2)
Upper1E <- assign1E$fStar + 1.96 * sqrt(diag(assign1E$cov_fStar))
Lower1E <- assign1E$fStar - 1.96 * sqrt(diag(assign1E$cov_fStar))

plotD <- data.frame(fStar=assign1E$fStar, xStar=assign1E$xStar, Lower=Lower1E, Upper=Upper1E)
xy <- data.frame(x=c(0.8, 0.4, -0.2, -0.6, -1), y=c(-0.664, 0.719, -0.94, -0.044, 0.768))
ggplot(plotD, aes(y=fStar, x=xStar)) + geom_point(col="darkorange") + ylim(-2,2.2) + 
  geom_line(data=plotD, aes(xStar, Lower), col="seagreen", size=1.1) +
  geom_line(data=plotD, aes(xStar, Upper), col="seagreen", size=1.1) +geom_point(data=xy, aes(x,y), size=3) +theme_classic()
```
  
  Compared to the plot in *d)*, the probability bands obtained with $\iota$ set to 1 looks much smoother and lies much closer to the curve for the posterior mean of *f*. Another change is that curve for the posterior mean of *f* no longer goes through all of the observed values. Instead the fitted curve appears to be an average between the observed values that lies closest to each other. With $\iota$ set to 1 the curve becomes all too smooth.    
  
  
#### Summarize - Assignment 1
As the number of observations increases we become more sure about the values of the posterior mean. How sure we become depends on the prior hyperparameters. $\sigma_f$ controls the uncertainty for the prior and length scale controls how many observations that affects the estimated posterior mean. 


### Assignment 2

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Assignment 2
library(kernlab)
temps <- read.csv("https://github.com/STIMALiU/AdvMLCourse/raw/master/GaussianProcess/Code/TempTullinge.csv",header=TRUE,   sep=";")
temps$time <- 1:2190
temps$day <- rep(1:365, 6)
temps$index <- rep(1:5, 438)
thinTemps <- subset(temps, temps$index == 1)[,1:4]
```

#### a)

The code where we define our own square exponential kernel function can be seen below. 
```{r}
sqExpK <- function(sigmaf = 1, ell = 1) {
  rval <- function(x, xStar = NULL) {
    return(sigmaf * exp (-1/2 * ((x-xStar)/ell)^2 ))
  }
  class(rval) <- "kernel"
  return(rval)
} 
```

When evaluated in the point *x* = 1 and *x'* = 2 the following value is given.
```{r, echo=TRUE}
sqExpKFunc = sqExpK(sigmaf = 1, ell = 2) # sqExpKFunc is a kernel FUNCTION
sqExpKFunc(1,2) # Evaluating the kernel in x=1, x'=2
```

The *kernelMatrix* is tested for the vectors **x** = ${(1,3,4)}^T$ and $x_*$ = ${(2,3,4)}^T$. Returned by this function is the covariance matrix **K**($x,x_*$). This matrix can be interpreted as following:  
The covariance between (x=1, x'=4) is 0.32. 
```{r, echo=TRUE}
# Computing the whole covariance matrix K from the kernel.
x <- as.matrix(c(1,3,4))
xStar <- as.matrix(c(2,3,4))
K <- kernelMatrix(kernel = sqExpKFunc, x = x, y = xStar) # So this is K(X,Xstar)
row.names(K) <- c(1,3,4)
colnames(K) <- c(2,3,4)
K
```

#### b)

The gaussian process regression model where the temperature is a function of time is estimated. At every data point in the training set is the posterior mean computed. In the graph below are the points in the data plotted together with the posterior mean.
The prior for $\sigma_n^2$ is 66.85 which is the variance obtained when a simple quadratic regression is estimated. The prior Hyperparameters $\sigma_f$ and $\iota$ are set to 20 and 0.2. 

```{r, echo=TRUE, fig.height=3.25, fig.width=7}
quadLM <- lm(temp ~ time + time^2, data=temps)
sigma_n <- var(quadLM$residuals)


sqExpKFunc = sqExpK(sigmaf = 20^2, ell = 0.2)
reg <- gausspr(thinTemps$time, thinTemps$temp, kernel = sqExpKFunc, var=sigma_n)
postMean <- data.frame(pred=predict(reg), y=thinTemps$temp, x=thinTemps$time)
plotB <- ggplot(postMean, aes(x=x, y=y)) + geom_point(col="grey75") + geom_line(data=postMean, aes(x=x, y=pred), col="darkorange", size=1.1) + theme_classic()
plotB
```

The orange curve are the predicted values and the grey dots the observed temperature. Overall, the fitted model seem to be rather good. There is some yearly variation which is modeled quite well. 


#### c)
The algorithm implemented in *1.a)* that was used for computing the posterior distribution of *f* is now used for computing the posterior variance of the results presented in *2.b)*. The result is presented by adding 95 % posterior probability bands to the plot of the posterior means presented in the previous exercise.

```{r, echo=TRUE, fig.height=3.25, fig.width=7}
covar_f <- PosteriorGP(x=thinTemps$time, y = thinTemps$temp, xStar = thinTemps$time, 
                       hyperParam=c(20^2, 0.2),sigmaNoise =sigma_n )
probBands <- data.frame(Upper=postMean[,1] + 1.96 * sqrt(diag(covar_f$cov_fStar)) , Lower = postMean[,1] - 1.96 * sqrt(diag(covar_f$cov_fStar)), x=thinTemps$time)
plotB + geom_line(data=probBands, aes(x=x, y=Upper), size=1.1, col="seagreen") + geom_line(data=probBands, aes(x=x, y=Lower), size=1.1, col="seagreen")
```

The 95 % posterior probability bands looks reasonable as all the observed temperatures lies inside the bands. However, the bands are perhaps a bit too wide. For example, the 95 % interval for the temperature during summer goes from around 0 to 35. 

#### d)

A model with temperature as a function of day is estimated. The posterior mean of this model is compared to the posterior mean of the model which had temperature as a function of time. 

```{r, echo=TRUE, fig.height=3.25, fig.width=7}
quadLM2 <- lm(temp ~ day + day^2, data=temps)
sigma_n2 <- var(quadLM2$residuals)
# Small difference, 67.3641 vs 64.10528
sqExpKFuncD = sqExpK(sigmaf = 20^2, ell = 1.2)
regD <- gausspr(thinTemps$day, thinTemps$temp, kernel = sqExpKFuncD, var=sigma_n2)
postMeanD <- data.frame(pred=predict(regD), y=thinTemps$temp, x=thinTemps$time)
postMT <- data.frame(rbind(postMean, postMeanD), model=c(rep("m1",438), rep("m2", 438)))
ggplot(postMT, aes(y=y, x=x)) + geom_point(col="grey75") + geom_line(aes(x=x, y=pred, col=model), size=1.1) +theme_classic() + scale_color_manual(values = c("darkorange", "seagreen"))
```

A comparsion of the curves gives that the model with day as input is constant and shows the exactly same pattern every year. Positive with this model is that it takes historical data into account when estimating the posterior mean. A negative aspect with this model is that it not allows any variation for the estimates given for specific day. 

The model that has time as input is more flexible. It allows the posterior mean for a day to differ between years as the estimate mainly is affected by the temperatures observed the weeks or days before during that year. The observed values the years before does not have any larger impact on the estimate. 

You could say that the pro for one model is a con for the second, and vice versa. 

#### e)

A periodic kernel is implemented and the formula for the kernel is given below.
$$K(x, x') = \sigma_f^2 \times exp(\frac{2sin^2(\pi |x-x'|/d)}{\iota_1^2}) \times exp(-0.5 \times(\frac{x-x'}{\iota_1^2})^2)$$
Two new parameters are then introduced. The correlation between the same day in different years is controlled by the parameter $\iota_2^2$ and the period is specified with the parameter *d*. 
A Gaussian process model is estimated with time as explanatory variable. 

```{r, echo=TRUE, warning=FALSE, message=FALSE, fig.height=4.25, fig.width=9}
PeriodicK <- function(sigmaf = 1, ell_one = 1, ell_two=1, d=1){
  rval <- function(x, xStar = NULL) {
    return(sigmaf * exp(-(2*sin(pi*abs(x-xStar)/d)^2)/ ell_one^2) *
             exp(-0.5 * (abs(x-xStar)^2 / ell_two^2)))
  }
  class(rval) <- "kernel"
  return(rval)
}
PeriodicKFunc = PeriodicK(sigmaf = 20^2, ell_one = 1, ell_two=10, d=365/sd(thinTemps$time))
#PeriodicKFunc is a kernel FUNCTION
regE <- gausspr(thinTemps$time, thinTemps$temp, kernel = PeriodicKFunc, var=sigma_n)
postMeanE <- data.frame(pred=predict(regE), y=thinTemps$temp, x=thinTemps$time)
plotE <- ggplot(postMeanE, aes(x=x, y=y)) + geom_point(col="grey75") + geom_line(data=postMeanE, aes(x=x, y=pred), col="darkorange", size=1.1) + theme_classic()
postMT <- rbind(postMT, data.frame(postMeanE, model="m3"))
plotAll <- ggplot(postMT, aes(x=x, y=pred)) + geom_line(aes(col=model, linetype=model),size=1.1, alpha=0.8) + theme_classic() + scale_color_manual(values = c("darkorange", "seagreen", "skyblue")) + scale_linetype_manual(values = c("dashed", "longdash", "solid")) + theme(legend.justification = c(1,0), legend.position=c(1,0)) + scale_y_continuous(limits = c(-17,20))
library(gridExtra)
grid.arrange(plotE, plotAll, ncol=2)
```

The periodic kernel combines both data from earlier years and data from days closer to the actual day. In practice it works as a combination of the models discussed in *2.d)*. The fitted curve in the plot to the left is similar to the earlier ones, with some exceptions. The posterior mean for the temperature during the winter increases for every year and the temperature for the summer looks to be constant. 

This result very well shows the ability of the periodic kernel. For some parts of the year the posterior mean is constant and for some it is more flexible. 


### Assignment 3

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Assignment 3
library(AtmRay)
data <- read.csv("https://github.com/STIMALiU/AdvMLCourse/raw/master/GaussianProcess/Code/banknoteFraud.csv",
                  header=FALSE, sep=",")
names(data) <- c("varWave","skewWave","kurtWave","entropyWave","fraud")
data[,5] <- as.factor(data[,5])
set.seed(111); SelectTraining <- sample(1:dim(data)[1], size = 1000, replace = FALSE)
Train <- data[SelectTraining, ]
Test <- data[-SelectTraining, ]
```

#### a)

```{r, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
# a)
cfA <- gausspr(fraud ~ varWave + skewWave, data=Train)
gridX <- seq(min(Train$varWave), max(Train$varWave), length=100)
gridY <- seq(min(Train$skewWave), max(Train$skewWave), length=100)
gridP <- meshgrid(gridX, gridY)
gridP <- cbind(c(gridP$x), c(gridP$y))
gridP <- data.frame(gridP)
names(gridP) <- names(Train)[1:2]
```

A model with the variables *varWave* and *skewWave* as input variables and fraud as target variable is estimated. The contours of the prediction probability for an observation to get classified as fraud or not is plotted below. In the contour plot is also training data added where fraud=1 are coloured blue and observations with fraud=0 coloured red. 

```{r, echo=TRUE}
probPredsA <- predict(cfA, gridP, type="probabilities")
contour(x=gridX, y= gridY,z=t(matrix(probPredsA[,2],100)),20)
points(Train[Train[,5]== 1,1],Train[Train[,5]== 1,2],col="blue")
points(Train[Train[,5]== 0,1],Train[Train[,5]== 0,2],col="red")
```

It can be seen that the model relatively well separates the training data correct with the prediction probabilities obtained with the model. This conclusion is also confirmed by the confusion matrix for the classifier on the training data and the accuracy.  

```{r, echo=TRUE}
# predict on the training set
confMatA <- table(predict(cfA,Train[,1:2]), Train[,5]) # confusion matrix
confMatA
cat("Accuracy:", sum(diag(confMatA)) / sum(confMatA))
```


#### b)
The estimated model is then used for making predictions on the test set and the results are presented in a confusion matrix. The accuracy for the test data is almost the same as for the training data. In order words does the model seem to be a rather good fit. 

```{r, echo=TRUE}
confMatB <- table(predict(cfA,Test[,1:2]), Test[,5]) # confusion matrix
confMatB
cat("Accuracy:", sum(diag(confMatB)) / sum(confMatB))
```


#### c)

A new model with two more input variables, *kurtWave* and *entropyWave*, is fitted. The confusion matrix and the accuracy of this model when the observations in the test data are predicted is presented below. A higher accuracy is recieved as all but one of the observations are classified correctly with this model.  

```{r, echo=TRUE, results='hide'}
cfC <- gausspr(fraud ~ varWave + skewWave + kurtWave + entropyWave, data=Train)
```

```{r, echo=FALSE}
confMatC <- table(predict(cfC,Test[,1:4]), Test[,5]) # confusion matrix
cat("Accuracy:", sum(diag(confMatC)) / sum(confMatC))
```




\pagebreak


# Lab 4

##Individual/Group report
### Assignment 1

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dlm)
library(gridExtra)
library(mvtnorm)
```


#### a) - Simulate the model

$y_t = Cx_t + \epsilon, \qquad \qquad \qquad \:\:\:\:\epsilon_t \sim  \mathcal{N}(0, \Omega_\epsilon)$  
$x_t = Ax_{t-1} + Bu_t + v_t, \qquad v_t \sim  \mathcal{N}(0, \Omega_v) \qquad x_o \sim  \mathcal{N}(\mu_0, \Sigma_0)$

Set $x_0=0$ and simulate T=100 observations from the following model. 

$y_t = 0.93x_t + \epsilon, \qquad \epsilon \sim  \mathcal{N}(0, \Omega_\epsilon=0.5)$  
$x_t = x_{t-1} + v_t, \qquad v_t \sim  \mathcal{N}(0, \Omega_v=0.1) \quad x_o \sim  \mathcal{N}(\mu_0=0, \Sigma_0=10)$   

A linear Guassian state space model is implemented in line with the given instructions. 100 observations are simulated from this model and the obtained states(x) and observations (y) are plotted in figure 1. 

```{r, echo=TRUE, fig.height=3, fig.width=7}
# Assignment 1

# a) - Simulate the model
ssm1 <- function(C=matrix(0), A=matrix(1), B=matrix(0), u=matrix(0), x0=0,
                 omegaE=0, omegaV=0, TimeS=0) {
  xt <- 0
  yt <- 0
  vt <- rnorm(1, 0, omegaV)
  xt[1] <- A%*%x0 + vt
  et <- rnorm(1, 0, omegaE)
  yt[1] <- C%*%xt + B%*%u + et
  for(i in 2:TimeS){
    vt <- rnorm(1, 0, omegaV)
    xt[i] <- A%*%xt[i-1] + vt
    et <- rnorm(1, 0, omegaE)
    yt[i] <- C%*%xt[i]+ B%*%u + et
  }
  res <- data.frame(Val=c(xt,yt), t= rep(1:TimeS,2), 
                    Value=rep(c("State","Observation"), each=TimeS))
  return(res)
}
set.seed(1234)
resA <- ssm1(C = matrix(0.93), x0 = 0, omegaE = sqrt(0.5), omegaV = sqrt(0.1), TimeS = 100)
plotA <- ggplot(resA, aes(x=t, y=Val, col=Value)) + geom_line(size=1) + theme_classic() + 
  ylim(-4,4) + scale_color_manual(values=c("skyblue", "royalblue")) 
plotA + labs(caption = "Figure 1")
```


#### b) - Filtering: Sequential state inference

The point estimates for the Kalman filter and the 0.95 probability bands for the point estimates are plotted together with the states and observations simulated in *a)*.

```{r, echo=TRUE, fig.height=3, fig.width=7, message=FALSE, warning=FALSE}
KalmanF <- function(yt, C=matrix(0), A=matrix(1),
                    B=matrix(0), u=matrix(0), omegaE=0, omegaV=0, TimeS=0){
  KalmanMat <- matrix(c(0,0), ncol=2, nrow=101)
  for(i in 1:TimeS){
    # Prediction update
    muBar_t <- A %*%  KalmanMat[i,1] + B%*%u
    sigmaBar_t <- A%*%KalmanMat[i,2]%*%t(A) + omegaV
    # Measurement update
    K_t <- sigmaBar_t%*%t(C) %*% solve(C%*%sigmaBar_t%*%t(C) + omegaE)
    KalmanMat[i+1,1] <- muBar_t + K_t %*% (yt[i] - C%*%muBar_t)
    KalmanMat[i+1,2] <- (diag(ncol(K_t))-K_t%*%C) %*% sigmaBar_t
  }
  KalmanFrame <- data.frame(KalmanMat[-1, ], Time=1:TimeS)
  return(KalmanFrame)
}

resB <- KalmanF(yt = resA[101:200,1], C = matrix(0.93), omegaE = 0.5, omegaV = 0.1, TimeS=100)
resB_1 <- data.frame(Val=c(resB[,1],resB[,1]+1.96*sqrt(resB[,2]),resB[,1]-1.96*sqrt(resB[,2])), Time=rep(1:100, 3), Value=rep(c("Kalman mean", "Upper", "Lower"), each=100))
                     
plotA + geom_line(data=resB_1, aes(x=Time, y=Val, col=Value), size=1) +ylim(-5,4)+
  scale_colour_manual(values=c("red3", "darkorange","skyblue","royalblue","darkorange"))+
  labs(x="Time", title="Kalman filter estimates with 0.95 probability intervals", caption="Figure 2")
```

In figure 3 are the Kalman filter estimates and probability intervals obtained by the implemented Kalman algorithm compared to the estimates returned by the *dlmFilter* function from the *dlm* package.   

```{r, echo=TRUE, fig.height=3, fig.width=7}
filteringB <- dlmFilter(y=resA[101:200,1], dlm(m0=0, V=0.5, W=0.1, C0=10, FF=0.93, GG=1) )

filterB <- data.frame(y=(filteringB$m)[-1])
filterVar <- unlist(dlmSvd2var(u = filteringB$U.C, d = filteringB$D.C))
filterB[101:300,1] <- c(filterB$y + 1.96*sqrt(filterVar[-1]), 
                        filterB$y - 1.96*sqrt(filterVar[-1]))
filterB$Time <- rep(1:100, 3)
filterB$Value <- rep(c("dlm Kalman mean", "dlm Upper", "dlm Lower"), each=100)

ggplot(resB_1, aes(x=Time, y=Val, col=Value,linetype=Value)) + geom_line(size=1) + theme_classic()+ ylim(-5,4)+ geom_line(data=filterB, aes(x=Time, y=y, col=Value, linetype=Value),size=1) +scale_colour_manual(values=c("purple","seagreen","seagreen","red3","darkorange","darkorange"))+scale_linetype_manual(values=c("dashed","dashed","dashed","solid","solid","solid"))+labs(title="Comparsion between dlmFilter and KalmanF function", caption="Figure 3")
```

Except for the first values are the point estimates and the probability intervals identical. 

#### c) - Prediction of state and data by simulation

When simulating we use *k* equals 5 and *i* goes from 1:1000. 
$$x_{T+k}^{(i)} \sim p(x_{T+k} \mid Y_{1:T})$$
$$y_{T+k}^{(i)} \sim p(y_{T+k} \mid x_{T+k}^{(i)})$$

The probability intervals for 5 future states and observations are plotted in figure 2 together with the data plotted in figure 2. The 5 future time steps are simulated 10 000 times and a 0.95 probabilty interval is given by the quantiles from the sample. 

```{r, echo=TRUE, fig.height=3, fig.width=7}
PredFunc <- function(C=matrix(0), A=matrix(1), B=matrix(0), u=matrix(0), x0=0,
                     omegaE=0, omegaV=0, TimeS=0){
  xt <- 0
  yt <- 0
  vt <- rnorm(1, 0, omegaV)
  xt[1] <- x0 + vt
  et <- rnorm(1, 0, omegaE)
  yt[1] <- C%*%xt + B%*%u + et
  for(i in 2:TimeS){
    vt <- rnorm(1, 0, omegaV)
    xt[i] <- A%*%xt[i-1] + vt
    et <- rnorm(1, 0, omegaE)
    yt[i] <- C%*%xt[i]+ B%*%u + et
  }
  res <- data.frame(x=c(xt, y=yt))
  return(res)
}

PredC <- as.data.frame(matrix(nrow=10,ncol=10000))
for(i in 1:10000){
  PredC[,i] <- PredFunc(C = matrix(0.93), x0 = rnorm(1, mean=resB$X1[100], 
                        sd=sqrt(resB$X2[100])),omegaE = sqrt(0.5), omegaV = sqrt(0.1),
                        TimeS = 5)
}
quanTs <- t(apply(PredC, 1, quantile, probs=c(0.025, 0.975)))
quanTsFrame <- data.frame(x=c(quanTs[,1], quanTs[,2]), Time= 100+rep(1:5, 4),
                 Var=rep(c("x", "y"), 2,each=5), InterV=rep(c("Lower", "Upper"), each=10))
quanTsFrame$Interval <- interaction(quanTsFrame$Var, quanTsFrame$InterV)
quanTsFrame$Interval<- factor(quanTsFrame$Interval, levels=rev(levels(quanTsFrame$Interval)))

ggplot(resA, aes(x=t, y=Val, col=Value)) + geom_line(size=1) + theme_classic() + ylim(-5,4) +
  geom_line(data=resB_1[1:100,], aes(x=Time, y=Val), col="seagreen", size=1.1) +
  geom_line(data=resB_1[101:200,], aes(x=Time, y=Val), col="darkorange", size=1) +
  geom_line(data=resB_1[201:300,], aes(x=Time, y=Val), col="darkorange", size=1) +
  geom_line(data=quanTsFrame, aes(x=Time, y=x, col=Interval), size=1, linetype="dashed") +
  scale_color_manual(values=c("skyblue", "royalblue","darkmagenta","darkmagenta","red3", "red3"))+ labs(x="Time", title="0.95 probability intervals for k=5", caption="Figure 4")
```

We notice how the width of these bands increase with time. This implies that we become more unsure for more distant future time steps.

#### d) - State and model parameter inference

Given the information from the lab instructions we create our FFBS function in order to be able to update the x values.
We implement a bayesian linear regression to be able to update our $\theta$ values. This will be used in our gibbs function and the update process will continue iteratively for 1000 iterations. These iterations can be seen below. It seems to converged rather quickly.

```{r, echo=TRUE, fig.height=3, fig.width=7}
KalmanF_2 <- function(yt, C=matrix(0), A=matrix(1),
                    B=matrix(0), u=matrix(0), omegaE=0, omegaV=0, TimeS=0){
  KalmanMat <- matrix(c(0,0,0,0), ncol=4, nrow=101)
  for(i in 1:TimeS){
    # Prediction update
    KalmanMat[i+1,3] <- A %*%  KalmanMat[i,1] + B%*%u #muBar_t 
    KalmanMat[i+1,4] <- A%*%KalmanMat[i,2]%*%t(A) + omegaV #sigmaBar_t
    # Measurement update
    K_t <- KalmanMat[i+1,4] %*%t(C) %*% solve(C%*%KalmanMat[i+1,4] %*%t(C) + omegaE)
    KalmanMat[i+1,1] <- KalmanMat[i+1,3] + K_t %*% (yt[i] - C%*%KalmanMat[i+1,3]) #mu_t
    KalmanMat[i+1,2] <- (diag(ncol(K_t))-K_t%*%C) %*% KalmanMat[i+1,4]  # sigma_t
  }
  KalmanFrame <- data.frame(KalmanMat[-1,], Time=1:TimeS)
  return(KalmanFrame)
}

BS <- function(A, Kalman, x){
  x_t <- data.frame(Val=0)
  for(k in (nrow(Kalman)-1):1){
    ht <- Kalman[k, 1] + Kalman[k, 2]%*%t(A)%*%solve(Kalman[k+1, 4]) %*% (x[k+1] - Kalman[k+1, 3])
    Ht <- Kalman[k,2]-Kalman[k,2]%*%t(A)%*%solve(Kalman[k+1,4])%*%A%*%Kalman[k,2]
    x_t[k,1] <- rnorm(1, mean=ht, sd=sqrt(Ht))
  }
  return(x_t)
}

bayesReg <- function(beta0, Xt, y, omega0, cov0){
  betaHat <- solve(t(Xt)%*%Xt)%*% t(Xt) %*% as.matrix(y)
  sigma0 <- diag(omega0/cov0,2)
  betaNew <- solve(t(Xt)%*%Xt + sigma0) %*% (t(Xt)%*%Xt%*%betaHat + sigma0%*%beta0)
  sigmaNew <- t(Xt)%*%Xt + sigma0
  Vari <- omega0 * solve(sigmaNew)
  coef <- rmvnorm(n=length(y), mean=betaNew, sigma=sqrt(diag(sqrt(diag(Vari)),length(betaNew))))
  return(coef)
}

# Gibbs sampling
y = resA[101:200, 1]
x = resA[1:100, 1]
set.seed(311015)
theta <- data.frame(matrix(ncol=2, nrow=1000)) #nrow sets number of sims
theta[1,2] <- 1  #start value for theta
Xt= matrix(data = c(rep(1,100),resA[1:100,1]), ncol=2)
for(j in 1:nrow(theta)){
  KalmanRes <- KalmanF_2(yt = y, C = matrix(theta[j,2]), omegaE = 0.5, omegaV = 0.1,
                         TimeS = 100)
  NewX <- (BS(A = matrix(1), Kalman = KalmanRes, x=x))
  Xt[1:99,2] <- as.numeric(NewX[,1])
  Xt[100,2] <- KalmanRes[100,1]
  coefs <- bayesReg(beta0 = matrix(data=c(0,0), ncol=1),Xt=Xt, y=y, omega0=0.5,cov0=100)
  theta[j+1,] <- colMeans(coefs) 
}
ggplot(theta, aes(y=X2, x=1:nrow(theta))) + geom_line() + theme_classic() + labs(caption="Figure 5", title="Trace plot", x="1:1000")
ggplot(theta, aes(x=X2)) + geom_density() +  theme_classic() +
  geom_vline(xintercept = c(mean(theta$X2),0.93), col=c("red","blue")) + labs(caption="Figure 6", title=expression(Posterior~distribution~of~theta)) 
```

In the density plot is the probability for different theta values plotted. The mean of our theta equals to $`r mean(theta[,2])`$ which is close to our true value. The true value, 0.93, could be obtained but the probability for this to happen is low.


### Assignment 2


#### a) - Estimating the variance components by MLE.

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Assignment 2
load("C:\\Users\\Gustav\\Documents\\AdvML\\Lab4\\CAPM.Rda")

# a) - Estimating the variance components by MLE

CAPMv <- CAPM_data[,c(10,17,18)]
Zt <- matrix(CAPM_data$MARKET - CAPM_data$RKFREE)
Yt <- matrix(CAPMv$IBM - CAPMv$RKFREE)

buildLocalTrend <- function(x,data){
  V = exp(x[1])
  W = diag(exp(x[2:3]),2,2)
  return(dlm(
    m0 = c(0,1),
    C0 = diag(c(100,0.5),2),
    FF = matrix(c(1,1),1,2),     
    GG = diag(2),
    JFF = matrix(c(0,1),1,2),
    V = V,
    W = W,
    X=data))
}

initVal <- c(1,1,1) # Initial values for optim on the estimated parameters 
MLEs <- dlmMLE(Yt, parm = initVal, build = buildLocalTrend, data = Zt)
dlmWithMLEs <- buildLocalTrend(MLEs$par, data = Zt)

```

By MLE the following estimates are given for the variance components. 

$\sigma_\epsilon^2$ = $`r (as.numeric(dlmWithMLEs[4]))`$,

$\sigma_{v^{(1)}}^2$ = $`r (data.frame((dlmWithMLEs[6]))[1,1])`$,

$\sigma_{v^{(2)}}^2$ = $`r (data.frame((dlmWithMLEs[6]))[2,2])`$



#### b) - Filtering and smoothing

The filtering and smoothing distribution is computed for $\alpha$ and $\beta$. In the figures below are these distributions plotted together with 0.95 probability intervals. 

The respective distributions for the alpha parameter are shown in the figures below. As expected, the same pattern is seen in both figures and a more smooth curve is returned for the smoothig distribution. 

```{r, echo=TRUE, fig.height=3, fig.width=7}
## Filtering ##
Filter_2b <- dlmFilter(y = Yt, dlmWithMLEs)

# Variances
u1 <- matrix((t(data.frame(Filter_2b$U.C)))[c(seq(1,242,2)),1])
u2 <- matrix((t(data.frame(Filter_2b$U.C)))[c(seq(2,242,2)),2])
u_1 <-list()
u_2 <- list()
for(i in 1:120){
  u_1[i]<-c(u1[i+1])
  u_2[i]<-c(u2[i+1])
}
Filter_y <- data.frame(alpha=c(Filter_2b$m)[2:121],beta=c(Filter_2b$m)[123:242])
Filter_y$alphaVar <- unlist(dlmSvd2var(u = u_1, d = matrix(Filter_2b$D.C[-1,1])))
Filter_y$betaVar <- unlist(dlmSvd2var(u = u_2, d = matrix(Filter_2b$D.C[-1,2])))
# Intervals
Filter_y$alphaLower <- Filter_y$alpha - 1.96*sqrt(Filter_y$alphaVar)
Filter_y$alphaUpper <- Filter_y$alpha + 1.96*sqrt(Filter_y$alphaVar)
Filter_y$betaLower <- Filter_y$beta - 1.96*sqrt(Filter_y$betaVar)
Filter_y$betaUpper <- Filter_y$beta + 1.96*sqrt(Filter_y$betaVar)

alphaFilter <- ggplot(Filter_y, aes(x=1:120, y=alpha)) + geom_line(col="seagreen", size=1) + theme_classic() + labs(title="Alpha filtered", caption="Figure 7") +
  geom_line(data=Filter_y, aes(x=1:120, y=alphaLower), col="darkorange", size=1) +
  geom_line(data=Filter_y, aes(x=1:120, y=alphaUpper), col="darkorange", size=1) 

betaFilter <- ggplot(Filter_y, aes(x=1:120, y=beta)) + geom_line( col="seagreen", size=1) + theme_classic()+ labs(title="Beta filtered", caption="Figure 9") +
  geom_line(data=Filter_y, aes(x=1:120, y=betaLower), col="darkorange", size=1) +
  geom_line(data=Filter_y, aes(x=1:120, y=betaUpper), col="darkorange", size=1)

#### Smoothed ####
Smooth_2b <- dlmSmooth(y = Yt, dlmWithMLEs)
Smooth_frame <- data.frame(alpha=Smooth_2b$s[-1,1], beta=Smooth_2b$s[-1,2])

covList = dlmSvd2var(Smooth_2b$U.S, Smooth_2b$D.S)
varMat = t(sapply(covList, FUN=function(x) sqrt(diag(x))))

Smooth_frame$alphaSd <- varMat[-1,1]
Smooth_frame$betaSd <- varMat[-1,2]
# Intervals
Smooth_frame$alphaLower <- Smooth_frame$alpha - 1.96*Smooth_frame$alphaSd
Smooth_frame$alphaUpper <- Smooth_frame$alpha + 1.96*Smooth_frame$alphaSd
Smooth_frame$betaLower <- Smooth_frame$beta - 1.96*Smooth_frame$betaSd
Smooth_frame$betaUpper <- Smooth_frame$beta + 1.96*Smooth_frame$betaSd

alphaSmooth <- ggplot(Smooth_frame, aes(x=1:120, y=alpha)) + geom_line(size=1, col="seagreen") + theme_classic() +labs(title="Alpha smoothed", caption="Figure 8") +
  geom_line(data=Smooth_frame, aes(x=1:120, y=alphaLower), size=1, col="darkorange") +
  geom_line(data=Smooth_frame, aes(x=1:120, y=alphaUpper), size=1, col="darkorange") 

betaSmooth <- ggplot(Smooth_frame, aes(x=1:120, y=beta)) + geom_line(size=1, col="seagreen")+theme_classic()+labs(title="Beta smoothed", caption="Figure 10") +
  geom_line(data=Smooth_frame, aes(x=1:120, y=betaLower),size=1 ,col="darkorange") +
  geom_line(data=Smooth_frame, aes(x=1:120, y=betaUpper),size=1, col="darkorange") 

grid.arrange(alphaFilter, alphaSmooth, ncol=2)
```

For the beta parameter is the pattern a little bit different for the respective distribution. In figure 10, for the smooth distribution, the value is increasing from around time point 50 and onwards. This pattern was not seen in the figure for the filter distribution. 

```{r, echo=FALSE, fig.height=3, fig.width=7}
grid.arrange(betaFilter, betaSmooth, ncol=2)
```


#### c) - Average sensitivity pre and post 1982 - a retrospective analysis

The retrospective analysis is performed by first computing the filtering distribution for the time points pre 1982. This is done by only using data for the first 48 observed time steps. Then, the same thing is done once again but for the data points from time point 49 to 120.

The *dlmBSample* function is used for making the *BS* step in the *FFBS* algorithm on the respective filtering distribution. With this approach are the mean values presented below obtained for the respective time period. Hence, it is concluded that the average sensitivity was higher after 82 than before 82. 

```{r, echo=TRUE, fig.height=3, fig.width=7}
filter_1 <- dlmFilter(Yt[1:48], buildLocalTrend(MLEs$par, Zt[1:48]))
filter_2 <- dlmFilter(Yt[49:120], buildLocalTrend(MLEs$par, Zt[49:120]))

set.seed(311015)
mean_1 <- mean(dlmBSample(filter_1)[,2])
mean_2 <- mean(dlmBSample(filter_2)[,2])

cat("Mean before 82:", mean_1)
cat("Mean after 82:", mean_2)
```


#### d) - Bayesian inference for the variance components


The first way is to use a likelihood that integrates out the states (i.e. will only be a function of the three sigma you have). This likelihood is then multiplied by a prior for the unknown's sigmas. This posterior can be sampled from using, for example, Metropolis-Hastings.

 See slide 6, Lecture 2 of how a likelihood a LGSS looks when it integrated the states. 



### Comments from Matias
När ni kör Gibbs sampling för att skatta theta så verkar det som att ni simulerar nya data y_1:T hela tiden. Det ska ni inte göra, om ni kollar på posteriorn så är ni intresserade av p(theta, x_1:T | y_1:T) - dvs BETINGAT på det data ni har (glöm inte att Bayesianer ALLTID betraktar data som känt. Det som är okänt i en bayesiansk analys är det vi försöker skatta). Resten av uppgift 1 d) har ni tänkt rätt på men åtgärda detta. Kör minst 5000 dragningar när ni samplar posteriorn om det går (koden blir snabbare när ni slipper simulera y hela tiden).  

Det ni är intresserade av är att beräkna sannolikheten
Pr( mean(beta_1:48) < mean(beta_49:100) |y_1:100). NOTERA att smoothing ger marginal fördelning för EN tidpunkt. Här är ni intresserade av simultan (joint) fördelning beta_1:100 | y_1:100. Drag från denna fördelning fås med FFBS. Det är alltså viktigt att ni kan skilja på vad det är FFBS ger (drag från en simultan fördelning givet y_1:T) och smoothing (drag från en marginal fördelning givet y_1:T)

2.d).  Det finns i princip två vägar att gå.

Första vägen är att använda en likelihood som integrerar ut states (dvs kommer bara vara en funktion av de tre sigma ni har). See slide 6, Lecture 2 för hur en likelihood in en LGSS ser ut när man integrerat ut states. Denna likelihood multipliceras sedan med en prior för de okända sigma's ni har, och denna posterior kan man sampla från med t.ex Metropolis-Hastings.

Den andra vägen är att använda Gibbs sampling på ett liknande sätt som i uppgift 1 d). I 1 d) så hade ni en lutning som var okänd och en varians som var känd - här är det tvärtom: ni kommer ha en okänd varians. Vi vet från Bayesian learning kursen att en normal-model med okänd varians har en conjugate prior som är Inv-scale-Chi2.



\pagebreak


# Example code (Lab 3 and Lab 4)
## R code for simulating from a GP
```{r, eval=FALSE}
# User input
nSim <- 10
sigmaF <- 0.1
l <- 2

#install.packages("mvtnorm")
library("mvtnorm")

# Setting up the kernel
SquaredExpKernel <- function(x1,x2,sigmaF=1,l=3){
  n1 <- length(x1)
  n2 <- length(x2)
  K <- matrix(NA,n1,n2)
  for (i in 1:n2){
    K[,i] <- sigmaF^2*exp(-0.5*( (x1-x2[i])/l)^2 )
  }
  return(K)
}


MeanFunc <- function(x){
  m <- sin(x)
  return(m)
}

SimGP <- function(m = 0,K,x,nSim,...){
  # Simulates nSim realizations (function) form a Gaussian process with mean m(x) and covariance K(x,x')
  # over a grid of inputs (x)
  n <- length(x)
  if (is.numeric(m)) meanVector <- rep(0,n) else meanVector <- m(x)
  covMat <- K(x,x,...)
  f <- rmvnorm(n, mean = meanVector, sigma = covMat)
  return(f)
}

xGrid <- seq(-5,5,length=20)
fSim <- SimGP(m=MeanFunc, K=SquaredExpKernel, x=xGrid, nSim, sigmaF, l)

plot(xGrid, fSim[1,], type="l", ylim = c(-3,3))
for (i in 2:nSim) {
  lines(xGrid, fSim[i,], type="l")
}
lines(xGrid,MeanFunc(xGrid), col = "red", lwd = 3)



# Plotting using manipulate package
library(manipulate)

plotGPPrior <- function(sigmaF, l, nSim){
  fSim <- SimGP(m=MeanFunc, K=SquaredExpKernel, x=xGrid, nSim, sigmaF, l)
  plot(xGrid, fSim[1,], type="l", ylim = c(-3,3), ylab="f(x)", xlab="x")
  for (i in 2:nSim) {
    lines(xGrid, fSim[i,], type="l")
  }
  lines(xGrid,MeanFunc(xGrid), col = "red", lwd = 3)
  title(paste('length scale =',l,', sigmaf =',sigmaF))
}

manipulate(
  plotGPPrior(sigmaF, l, nSim = 10),
  sigmaF = slider(0, 2, step=0.1, initial = 1, label = "SigmaF"),
  l = slider(0, 2, step=0.1, initial = 1, label = "Length scale, l")
)
```

## Quick demo of the R package kernlab
```{r, eval=FALSE}
##############################################################################################
## Kernlab demo for Gaussian processes regression and classfication
## Author: Mattias Villani, Linkoping University. http://mattiasvillani.com
##############################################################################################


#############################################################################
###    Prelims: Setting path and installing/loading packages            #####
#############################################################################
setwd('/Users/mv/Dropbox/Teaching/AdvML/GaussianProcess/Code')
#install.packages('kernlab')
#install.packages("AtmRay") # To make 2D grid like in Matlab's meshgrid
library(kernlab)
library(AtmRay)


###############################################
###    Messin' around with kernels        #####
###############################################
# This is just to test how one evaluates a kernel function
# and how one computes the covariance matrix from a kernel function
X <- matrix(rnorm(12), 4, 3)
Xstar <- matrix(rnorm(15), 5, 3)
ell <- 1
SEkernel <- rbfdot(sigma = 1/(2*ell^2)) # Note how I reparametrize the rbfdo (which is the SE kernel) in kernlab
SEkernel(1,2) # Just a test - evaluating the kernel in the points x=1 and x'=2
# Computing the whole covariance matrix K from the kernel. Just a test.
K <- kernelMatrix(kernel = SEkernel, x = X, y = Xstar) # So this is K(X,Xstar)

# Own implementation of Matern with nu = 3/2 (See RW book equation 4.17)
# Note that a call of the form kernelFunc <- Matern32(sigmaf = 1, ell = 0.1) returns a kernel FUNCTION.
# You can now evaluate the kernel at inputs: kernelFunc(x = 3, y = 4).
# Note also that class(kernelFunc) is of class "kernel", which is a class defined by kernlab.
Matern32 <- function(sigmaf = 1, ell = 1) 
{
  rval <- function(x, y = NULL) {
      r = sqrt(crossprod(x-y));
      return(sigmaf^2*(1+sqrt(3)*r/ell)*exp(-sqrt(3)*r/ell))
    }
  class(rval) <- "kernel"
  return(rval)
} 


# Testing our own defined kernel function.
X <- matrix(rnorm(12), 4, 3) # Simulating some data
Xstar <- matrix(rnorm(15), 5, 3)
MaternFunc = Matern32(sigmaf = 1, ell = 2) # MaternFunc is a kernel FUNCTION
MaternFunc(c(1,1),c(2,2)) # Evaluating the kernel in x=c(1,1), x'=c(2,2)
# Computing the whole covariance matrix K from the kernel.
K <- kernelMatrix(kernel = MaternFunc, x = X, y = Xstar) # So this is K(X,Xstar)


###############################################
###      Regression on the LIDAR data       ###
###############################################
lidarData <- read.table('https://raw.githubusercontent.com/STIMALiU/AdvMLCourse/master/GaussianProcess/Code/LidarData', 
                        header = T)
LogRatio <- lidarData$LogRatio
Distance <- lidarData$Distance

# Estimating the noise variance from a third degree polynomial fit
polyFit <- lm(LogRatio ~  Distance + I(Distance^2) + I(Distance^3) )
sigmaNoise = sd(polyFit$residuals)

plot(Distance,LogRatio)

# Fit the GP with built in Square expontial kernel (called rbfdot in kernlab)
ell <- 2
GPfit <- gausspr(Distance, LogRatio, kernel = rbfdot, kpar = list(sigma = 1/(2*ell^2)), var = sigmaNoise^2)
meanPred <- predict(GPfit, Distance) # Predicting the training data. To plot the fit.
lines(Distance, meanPred, col="blue", lwd = 2)

# Fit the GP with home made Matern
sigmaf <- 1
ell <- 2
# GPfit <- gausspr(Distance, LogRatio, kernel = Matern32(ell=1)) # NOTE: this also works and is the same as the next line.
GPfit <- gausspr(Distance, LogRatio, kernel = Matern32, kpar = list(sigmaf = sigmaf, ell=ell), var = sigmaNoise^2) 
meanPred <- predict(GPfit, Distance)
lines(Distance, meanPred, col="purple", lwd = 2)

# Trying another length scale
sigmaf <- 1
ell <- 1
GPfit <- gausspr(Distance, LogRatio, kernel = Matern32, kpar = list(sigmaf = sigmaf, ell=ell), var = sigmaNoise^2) 
meanPred <- predict(GPfit, Distance)
lines(Distance, meanPred, col="green", lwd = 2)

# And now with a different sigmaf
sigmaf <- 0.1
ell <- 2
GPfit <- gausspr(Distance, LogRatio, kernel = Matern32, kpar = list(sigmaf = sigmaf, ell=ell), var = sigmaNoise^2) 
meanPred <- predict(GPfit, Distance)
lines(Distance, meanPred, col="black", lwd = 2)



###############################################
####       Classification on Iris data      ###
###############################################
data(iris)
GPfitIris <- gausspr(Species ~  Sepal.Length + Sepal.Width + Petal.Length + Petal.Width, data=iris)
GPfitIris

# predict on the training set
predict(GPfitIris,iris[,1:4])
table(predict(GPfitIris,iris[,1:4]), iris[,5]) # confusion matrix

# Now using only Sepal.Length and Sepal.Width to classify
GPfitIris <- gausspr(Species ~  Sepal.Length + Sepal.Width, data=iris)
GPfitIris
# predict on the training set
predict(GPfitIris,iris[,1:2])
table(predict(GPfitIris,iris[,1:2]), iris[,5]) # confusion matrix

# Now using only  Petal.Length + Petal.Width to classify
GPfitIris <- gausspr(Species ~ Petal.Length + Petal.Width, data=iris)
GPfitIris
# predict on the training set
predict(GPfitIris,iris[,3:4])
table(predict(GPfitIris,iris[,3:4]), iris[,5]) # confusion matrix

# class probabilities 
probPreds <- predict(GPfitIris, iris[,3:4], type="probabilities")
x1 <- seq(min(iris[,3]),max(iris[,3]),length=100)
x2 <- seq(min(iris[,4]),max(iris[,4]),length=100)
gridPoints <- meshgrid(x1, x2)
gridPoints <- cbind(c(gridPoints$x), c(gridPoints$y))

gridPoints <- data.frame(gridPoints)
names(gridPoints) <- names(iris)[3:4]
probPreds <- predict(GPfitIris, gridPoints, type="probabilities")

# Plotting for Prob(setosa)
contour(x1,x2,matrix(probPreds[,1],100), 20, xlab = "Petal.Length", ylab = "Petal.Width", main = 'Prob(Setosa) - Setosa is red')
points(iris[iris[,5]=='setosa',3],iris[iris[,5]=='setosa',4],col="red")
points(iris[iris[,5]=='virginica',3],iris[iris[,5]=='virginica',4],col="blue")
points(iris[iris[,5]=='versicolor',3],iris[iris[,5]=='versicolor',4],col="green")

# Plotting for Prob(Versicolor)
contour(x1,x2,matrix(probPreds[,2],100), 20, xlab = "Petal.Length", ylab = "Petal.Width", main = 'Prob(Versicolor) - Versicolor is green')
points(iris[iris[,5]=='setosa',3],iris[iris[,5]=='setosa',4],col="red")
points(iris[iris[,5]=='virginica',3],iris[iris[,5]=='virginica',4],col="blue")
points(iris[iris[,5]=='versicolor',3],iris[iris[,5]=='versicolor',4],col="green")


# Plotting for Prob(virginica)
contour(x1,x2,matrix(probPreds[,3],100), 20, xlab = "Petal.Length", ylab = "Petal.Width", main = 'Prob(Virginica) - Virginica is blue')
points(iris[iris[,5]=='setosa',3],iris[iris[,5]=='setosa',4],col="red")
points(iris[iris[,5]=='virginica',3],iris[iris[,5]=='virginica',4],col="blue")
points(iris[iris[,5]=='versicolor',3],iris[iris[,5]=='versicolor',4],col="green")

# Plotting the decision boundaries
meanPred <- matrix(max.col(probPreds),100)
plot(gridPoints,  pch=".", cex=3, col=ifelse(meanPred==1, "red", ifelse(meanPred==2, "green", "blue")))
points(iris[iris[,5]=='setosa',3],iris[iris[,5]=='setosa',4],col="red", cex=10, pch=".")
points(iris[iris[,5]=='virginica',3],iris[iris[,5]=='virginica',4],col="blue",  cex=10, pch=".")
points(iris[iris[,5]=='versicolor',3],iris[iris[,5]=='versicolor',4],col="green",  cex=10, pch=".")

```

## UC model demo on Nile flow data 
```{r, eval=FALSE}
# Unobserved components model
# My own spin of the analysis in the vignette to the rucm package in R
# Mattias Villani, http://mattiasvillani.com


#install.packages('rucm')
library(rucm)

# Plotting the data
# The Nile dataset in is the datasets package
timeNile <- seq(1871, 1970)
nTraining <- length(timeNile) # Total number of data points
nPreds <- 20
par(cex.axis = 0.8, cex.main = 0.8, cex.lab = 0.8)
plot(Nile, ylab = "Flow of Nile", lwd = 2, xlab = 'year', 
     xlim = c(1,nTraining+nPreds), main = "Modeling the Nile flow")
lines(seq(1,nTraining), Nile, lwd = 2)

#### Fitting a local level model ###
# y(t)  = mu(t) + eps(t)   [sigma2_eps is the Irregular_Variance in rucm]
# mu(t) = mu(t-1) + nu(t)  [sigma2_nu is the Level_Variance in rucm]
localLevelModel <- ucm(formula = Nile~0, data = Nile, level = TRUE)
localLevelModel
lines(seq(1,nTraining), localLevelModel$s.level, col = "red", lwd = 2)

#### Fitting a local trend model ###
# y(t)  = mu(t) + eps(t)           [sigma2_eps is the Irregular_Variance in rucm]
# mu(t) = mu(t-1) + beta_t + nu(t) [sigma2_nu is the Level_Variance in rucm]
# beta(t) = beta(t-1) + eta(t)     [sigma2_eta is the Slope_Variance in rucm]
localTrendModel <- ucm(formula = Nile~0, data = Nile, level = TRUE, slope = TRUE)

# Adding the fit from the local trend model
#plot(Nile, ylab = "Flow of Nile", lwd = 2, xlab = 'year')
lines(seq(1,nTraining), localTrendModel$s.level + localTrendModel$s.slope, col = "blue", lwd = 2, xlab = 'year')

# Make predictions from local level model
predsLocalLevel <- predict(localLevelModel$model, n.ahead = nPreds)
predsLocalTrend <- predict(localTrendModel$model, n.ahead = nPreds)

lines(seq(nTraining + 1,nTraining + nPreds), predsLocalLevel, 
      type = "l", lwd = 2, col = "red", lty = 2)
lines(seq(nTraining + 1,nTraining + nPreds), predsLocalTrend, 
      type = "l", lwd = 2, col = "blue", lty = 2)
legend("topright", legend = c("Data","Fit local level", "Fit local trend",
                              "Predictions local level","Predictions local trend"), 
       col = c("black","red","blue","red","blue"), lty = c(1,1,1,2,2), lwd = 2, cex = 0.45)


# Plotting the level and trend components separately. 
plot(timeNile, localTrendModel$s.level, type = "l", col = "red", lwd = 2, 
     main = "Inferred level", xlab = 'year', ylab = "")
plot(timeNile, localTrendModel$s.slope, type = "l", col = "red", lwd = 2, 
     main = "Inferred slope", xlab = 'year', ylab = "")
```

## DLM package demo on Nile flow data
```{r, eval=FALSE}
# Demo of the dlm package using the nile data
# Author: Mattias Villani, http://mattiasvillani.com

#install.packages("dlm")
library(dlm)

# Plotting the data
# The Nile dataset in is the datasets package
timeNile <- seq(1871, 1970)
nTraining <- length(timeNile) # Total number of data points
nPreds <- 20
par(cex.axis = 0.8, cex.main = 0.8, cex.lab = 0.8)
plot(Nile, ylab = "Flow of Nile", lwd = 2, xlab = 'year', 
     xlim = c(1,nTraining+nPreds), main = "Modeling the Nile flow")
lines(seq(1,nTraining), Nile, lwd = 2)

# Set up the local trend model
# the state is theta_t = (mu_t, beta_)'
m0 = c(0,0)     # Prior mean for the first state theta_0
C0 = 100*diag(2) # Prior covariance for the first state theta_0
FF = matrix(c(1,0),1,2)     # The mapping from the states to the measurements. 
                            # Note FF must be a matrix object even if it is a vector ...
GG = matrix(c(1,1,0,1), 2, 2, byrow = T)
V = 10
W = 0.01*diag(2)

dlmLocalTrend <- dlm(m0 = m0,C0 = C0, FF = FF, GG = GG, V = V, W = W)

# Kalman filtering
dlmKFfilt <- dlmFilter(y = Nile, dlmLocalTrend)

# Plotting the sum of the filtered states
lines(seq(1,nTraining), dlmKFfilt$a[,1] + dlmKFfilt$a[,2], col = "red", lwd = 2)

# Plotting the predicted response, y
lines(seq(1,nTraining), dlmKFfilt$f, col = "green", lwd = 2)

# State smoothing
dlmSmoothed <- dlmSmooth(y = Nile, dlmLocalTrend)

# Plotting the data the sum of the filtered and smoothed series for the two states.
par(cex.axis = 0.8, cex.main = 0.8, cex.lab = 0.8)
plot(Nile, ylab = "Flow of Nile", lwd = 2, xlab = 'year', 
     xlim = c(1,nTraining+nPreds), main = "Modeling the Nile flow")
lines(seq(1,nTraining), Nile, lwd = 2)

# Plotting the sum of the filtered states
lines(seq(1,nTraining), dlmKFfilt$a[,1] + dlmKFfilt$a[,2], col = "red", lwd = 2)

# Plotting the sum of the smoothed states. Note that dlmSmoothed$s[1,] is theta at time t=0
lines(seq(1,nTraining), dlmSmoothed$s[2:(nTraining+1),1] + dlmSmoothed$s[2:(nTraining+1),2], col = "blue", lwd = 2)


# Parameter estimation

# First create so called build function that takes the parameters that you want to estimate
# as input and returns a list (or a dlm object) with the state space matrices as output
buildLocalTrend <- function(x){
  
  V = exp(x[1])   # NOTE THE EXPONENTIAL. x must be unrestricted parameters, otherwise optim will go bananas
  W = diag(exp(x[2:3]),2,2)
  
  return(dlm(
    m0 = rep(0,2),
    C0 = 100*diag(2),   
    FF = matrix(c(1,0),1,2),     
    GG = matrix(c(1,1,0,1), 2, 2, byrow = T),
    V = V,
    W = W))
  
}


initVal <- c(1,1,1) # Initial values for optim on the estimated parameters 
dlmLocalTrend <- buildLocalTrend(initVal) # Just to see that the build function works
MLEs <- dlmMLE(Nile, parm = initVal, build = buildLocalTrend)
dlmWithMLEs <- buildLocalTrend(MLEs$par)

# Let's plot the filtered and smoothed states using the MLEs

# First plot the Nile data
par(cex.axis = 0.8, cex.main = 0.8, cex.lab = 0.8)
plot(Nile, ylab = "Flow of Nile", lwd = 2, xlab = 'year', 
     xlim = c(1,nTraining+nPreds), main = "Modeling the Nile flow")
lines(seq(1,nTraining), Nile, lwd = 2)

# Filtering
dlmKFfilt <- dlmFilter(y = Nile, dlmWithMLEs)

# Plotting the sum of the filtered states
lines(seq(1,nTraining), dlmKFfilt$a[,1] + dlmKFfilt$a[,2], col = "red", lwd = 2)

# State smoothing
dlmSmoothed <- dlmSmooth(y = Nile, dlmWithMLEs)

# Plotting the sum of the smoothed states. Note that dlmSmoothed$s[1,] is theta at time t=0
lines(seq(1,nTraining), dlmSmoothed$s[2:(nTraining+1),1] 
      + dlmSmoothed$s[2:(nTraining+1),2], col = "blue", lwd = 2)


# Bayesian analysis
nDraws <- 1000
nStates <- 2
postDraws <- array(NA, c(nTraining, nDraws, nStates) )
for (i in 1:nDraws){
  postDraws[,i,] <- dlmBSample(dlmKFfilt)[-1,]
}

# Plotting the smoothed series (posterior mean)
# and the first 3 draws from posterior
par(cex.axis = 0.8, cex.main = 0.8, cex.lab = 0.8)
plot(Nile, ylab = "Flow of Nile", lwd = 2, xlab = 'year', 
     xlim = c(1,nTraining+nPreds), main = "Modeling the Nile flow")
lines(seq(1,nTraining), Nile, lwd = 2)
lines(seq(1,nTraining), dlmSmoothed$s[2:(nTraining+1),1] 
      + dlmSmoothed$s[2:(nTraining+1),2], col = "red", lwd = 2)
for (i in 1:3){
  lines(seq(1,nTraining), postDraws[,i,1] + postDraws[,i,2], col = "blue", lwd = 1)
}

# Plotting the marginal posterior of the state at time t
hist(postDraws[50,,1], 40)
```




\pagebreak


# Lab instructions
## Lab 1
###(1)  
Show that multiple runs hill-climbing algorithm can return non-equivalent DAGs. Can
we check if two DAGs are equivalent by checking if they receive the same score ?
Hint: Check the function hc in the bnlearn package. Note that you can specify the
initial structure as well as the number of random restarts. Use these options to answer
the question. You may also want to use the functions plot, arcs, vstructs, cpdag and
all.equal.

###(2)  
Show that increasing the equivalent sample size (a.k.a imaginary sample size) in the
BDeu score decreases regularization. Explain why this happens.
Hint: Run some structure learning algorithm (e.g. check the function hc in the bnlearn
package) multiple times and show that it tends to end in more densely connected
DAGs. Or produce a histogram of the scores of a random sample of DAGs with different
imaginary sample sizes and see if they come closer or not one to another (e.g. check
the functions hist, random.graph, sapply and score in the bnlearn and core packages).

###(3)  
You already know the LS algorithm for inference in BNs, which is an exact algorithm.
There are also approximate algorithms for when the exact ones are too demanding
computationally. Compare the answers given to some queries by exact and approximate
inference algorithms. When running the approximate algorithm several times,
why may we obtain different answers ? Is the approximate algorithm equally accurate
when the query includes no observed nodes and when it includes many observerd
nodes ?
Hint: For exact inference, you may need the functions bn.fit and as.grain from the
bnlearn package, and the functions compile, setFinding and querygrain from the package
gRain. For approximate inference, you may need the functions prop.table, table
and cpdist from the bnlearn package.

###(4)  
There are 29281 DAGs with five nodes. Compute approximately the fraction of the
29281 DAGs that represent different independence models. How does this affect structure
learning ?
Hint: You do not need to produce the 29281 DAGs. Instead you can sample DAGs
uniformly, convert each DAG in the sample to its essential graph (a.k.a completed
partial DAG or CPDAG), and then count how many different essential graphs you have
left. You can do all this with the functions random.graph, cpdag, lapply, unique and
length in the bnlearn and core packages.



## Lab 2
###(1)  
Build a HMM for the scenario described above.

###(2)  
Simulate the HMM for 100 time steps.


###(3)  
Discard the hidden states from the sample obtained above. Use the remaining observations
to compute the filtered and smoothed probability distributions for each of the
100 time points. Compute also the most probable path.


###(4)  
Compute the accuracy of the filtered and smoothed probability distributions, and of the
most probable path. That is, compute the percentage of the true hidden states that are
guessed by each method.
Hint: Note that the function forward in the HMM package returns probabilities in
log scale. You may need to use the functions exp and prop.table in order to obtain a
normalized probability distribution. You may also want to use the functions apply and
which.max to find out the most probable states. Finally, recall that you can compare
two vectors A and B elementwise as A==B, and that the function table will count the
number of times that the different elements in a vector occur in the vector.


###(5)  
Repeat the previous exercise with different simulated samples. In general, the smoothed
distributions should be more accurate than the filtered distributions. Why ? In general,
the smoothed distributions should be more accurate than the most probable path, too.
Why ?


###(6)  
Is it true that the more observations you have the better you know where the robot is ?
Hint: You may want to compute the entropy of the filtered distributions with the function
entropy.empirical of the package entropy.


###(7)  
Consider any of the samples above of length 100. Compute the probabilities of the
hidden states for the time step 101.





