---
title: "Lab 4 - Advanced Machine Learning"
author: "Gustav Sternelöv"
date: "October 11, 2016"
output: pdf_document
---

# Assignment 1

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dlm)
library(gridExtra)
library(mvtnorm)
```


## a) - Simulate the model

$y_t = Cx_t + \epsilon, \qquad \qquad \qquad \:\:\:\:\epsilon_t \sim  \mathcal{N}(0, \Omega_\epsilon)$  
$x_t = Ax_{t-1} + Bu_t + v_t, \qquad v_t \sim  \mathcal{N}(0, \Omega_v) \qquad x_o \sim  \mathcal{N}(\mu_0, \Sigma_0)$

Set $x_0=0$ and simulate T=100 observations from the following model. 

$y_t = 0.93x_t + \epsilon, \qquad \epsilon \sim  \mathcal{N}(0, \Omega_\epsilon=0.5)$  
$x_t = x_{t-1} + v_t, \qquad v_t \sim  \mathcal{N}(0, \Omega_v=0.1) \quad x_o \sim  \mathcal{N}(\mu_0=0, \Sigma_0=10)$   

A linear Guassian state space model is implemented in line with the given instructions. 100 observations are simulated from this model and the obtained states(x) and observations (y) are plotted in figure 1. 

```{r, echo=FALSE, fig.height=3, fig.width=7}
# Assignment 1

# a) - Simulate the model
ssm1 <- function(C=matrix(0), A=matrix(1), B=matrix(0), u=matrix(0), x0=0,
                 omegaE=0, omegaV=0, TimeS=0) {
  xt <- 0
  yt <- 0
  vt <- rnorm(1, 0, omegaV)
  xt[1] <- A%*%x0 + vt
  et <- rnorm(1, 0, omegaE)
  yt[1] <- C%*%xt + B%*%u + et
  for(i in 2:TimeS){
    vt <- rnorm(1, 0, omegaV)
    xt[i] <- A%*%xt[i-1] + vt
    et <- rnorm(1, 0, omegaE)
    yt[i] <- C%*%xt[i]+ B%*%u + et
  }
  res <- data.frame(Val=c(xt,yt), t= rep(1:TimeS,2), 
                    Value=rep(c("State","Observation"), each=TimeS))
  return(res)
}
set.seed(1234)
resA <- ssm1(C = matrix(0.93), x0 = 0, omegaE = sqrt(0.5), omegaV = sqrt(0.1), TimeS = 100)
plotA <- ggplot(resA, aes(x=t, y=Val, col=Value)) + geom_line(size=1) + theme_classic() + 
  ylim(-4,4) + scale_color_manual(values=c("skyblue", "royalblue")) 
plotA + labs(caption = "Figure 1")
```


## b) - Filtering: Sequential state inference

The point estimates for the Kalman filter and the 0.95 probability bands for the point estimates are plotted together with the states and observations simulated in *a)*.

```{r, echo=FALSE, fig.height=3, fig.width=7, message=FALSE, warning=FALSE}
KalmanF <- function(yt, C=matrix(0), A=matrix(1),
                    B=matrix(0), u=matrix(0), omegaE=0, omegaV=0, TimeS=0){
  KalmanMat <- matrix(c(0,0), ncol=2, nrow=101)
  for(i in 1:TimeS){
    # Prediction update
    muBar_t <- A %*%  KalmanMat[i,1] + B%*%u
    sigmaBar_t <- A%*%KalmanMat[i,2]%*%t(A) + omegaV
    # Measurement update
    K_t <- sigmaBar_t%*%t(C) %*% solve(C%*%sigmaBar_t%*%t(C) + omegaE)
    KalmanMat[i+1,1] <- muBar_t + K_t %*% (yt[i] - C%*%muBar_t)
    KalmanMat[i+1,2] <- (diag(ncol(K_t))-K_t%*%C) %*% sigmaBar_t
  }
  KalmanFrame <- data.frame(KalmanMat[-1, ], Time=1:TimeS)
  return(KalmanFrame)
}

resB <- KalmanF(yt = resA[101:200,1], C = matrix(0.93), omegaE = 0.5, omegaV = 0.1, TimeS=100)
resB_1 <- data.frame(Val=c(resB[,1],resB[,1]+1.96*sqrt(resB[,2]),resB[,1]-1.96*sqrt(resB[,2])), Time=rep(1:100, 3), Value=rep(c("Kalman mean", "Upper", "Lower"), each=100))
                     
plotA + geom_line(data=resB_1, aes(x=Time, y=Val, col=Value), size=1) +ylim(-5,4)+
  scale_colour_manual(values=c("red3", "darkorange","skyblue","royalblue","darkorange"))+
  labs(x="Time", title="Kalman filter estimates with 0.95 probability intervals", caption="Figure 2")
```

In figure 3 are the Kalman filter estimates and probability intervals obtained by the implemented Kalman algorithm compared to the estimates returned by the *dlmFilter* function from the *dlm* package.   

```{r, echo=FALSE, fig.height=3, fig.width=7}
filteringB <- dlmFilter(y=resA[101:200,1], dlm(m0=0, V=0.5, W=0.1, C0=10, FF=0.93, GG=1) )

filterB <- data.frame(y=(filteringB$m)[-1])
filterVar <- unlist(dlmSvd2var(u = filteringB$U.C, d = filteringB$D.C))
filterB[101:300,1] <- c(filterB$y + 1.96*sqrt(filterVar[-1]), 
                        filterB$y - 1.96*sqrt(filterVar[-1]))
filterB$Time <- rep(1:100, 3)
filterB$Value <- rep(c("dlm Kalman mean", "dlm Upper", "dlm Lower"), each=100)

ggplot(resB_1, aes(x=Time, y=Val, col=Value,linetype=Value)) + geom_line(size=1) + theme_classic()+ ylim(-5,4)+ geom_line(data=filterB, aes(x=Time, y=y, col=Value, linetype=Value),size=1) +scale_colour_manual(values=c("purple","seagreen","seagreen","red3","darkorange","darkorange"))+scale_linetype_manual(values=c("dashed","dashed","dashed","solid","solid","solid"))+labs(title="Comparsion between dlmFilter and KalmanF function", caption="Figure 3")
```

Except for the first values are the point estimates and the probability intervals identical. 

## c) - Prediction of state and data by simulation

When simulating we use *k* equals 5 and *i* goes from 1:1000. 
$$x_{T+k}^{(i)} \sim p(x_{T+k} \mid Y_{1:T})$$
$$y_{T+k}^{(i)} \sim p(y_{T+k} \mid x_{T+k}^{(i)})$$

The probability intervals for 5 future states and observations are plotted in figure 2 together with the data plotted in figure 2. The 5 future time steps are simulated 10 000 times and a 0.95 probabilty interval is given by the quantiles from the sample. 

```{r, echo=FALSE, fig.height=3, fig.width=7}
PredFunc <- function(C=matrix(0), A=matrix(1), B=matrix(0), u=matrix(0), x0=0,
                     omegaE=0, omegaV=0, TimeS=0){
  xt <- 0
  yt <- 0
  vt <- rnorm(1, 0, omegaV)
  xt[1] <- x0 + vt
  et <- rnorm(1, 0, omegaE)
  yt[1] <- C%*%xt + B%*%u + et
  for(i in 2:TimeS){
    vt <- rnorm(1, 0, omegaV)
    xt[i] <- A%*%xt[i-1] + vt
    et <- rnorm(1, 0, omegaE)
    yt[i] <- C%*%xt[i]+ B%*%u + et
  }
  res <- data.frame(x=c(xt, y=yt))
  return(res)
}

PredC <- as.data.frame(matrix(nrow=10,ncol=10000))
for(i in 1:10000){
  PredC[,i] <- PredFunc(C = matrix(0.93), x0 = rnorm(1, mean=resB$X1[100], 
                        sd=sqrt(resB$X2[100])),omegaE = sqrt(0.5), omegaV = sqrt(0.1),
                        TimeS = 5)
}
quanTs <- t(apply(PredC, 1, quantile, probs=c(0.025, 0.975)))
quanTsFrame <- data.frame(x=c(quanTs[,1], quanTs[,2]), Time= 100+rep(1:5, 4),
                 Var=rep(c("x", "y"), 2,each=5), InterV=rep(c("Lower", "Upper"), each=10))
quanTsFrame$Interval <- interaction(quanTsFrame$Var, quanTsFrame$InterV)
quanTsFrame$Interval<- factor(quanTsFrame$Interval, levels=rev(levels(quanTsFrame$Interval)))

ggplot(resA, aes(x=t, y=Val, col=Value)) + geom_line(size=1) + theme_classic() + ylim(-5,4) +
  geom_line(data=resB_1[1:100,], aes(x=Time, y=Val), col="seagreen", size=1.1) +
  geom_line(data=resB_1[101:200,], aes(x=Time, y=Val), col="darkorange", size=1) +
  geom_line(data=resB_1[201:300,], aes(x=Time, y=Val), col="darkorange", size=1) +
  geom_line(data=quanTsFrame, aes(x=Time, y=x, col=Interval), size=1, linetype="dashed") +
  scale_color_manual(values=c("skyblue", "royalblue","darkmagenta","darkmagenta","red3", "red3"))+ labs(x="Time", title="0.95 probability intervals for k=5", caption="Figure 4")
```

We notice how the width of these bands increase with time. This implies that we become more unsure for more distant future time steps.

## d) - State and model parameter inference

Given the information from the lab instructions we create our FFBS function in order to be able to update the x values.
We implement a bayesian linear regression to be able to update our $\theta$ values. This will be used in our gibbs function and the update process will continue iteratively for 1000 iterations. These iterations can be seen below. It seems to converged rather quickly.

```{r, echo=FALSE, fig.height=3, fig.width=7}
KalmanF_2 <- function(yt, C=matrix(0), A=matrix(1),
                    B=matrix(0), u=matrix(0), omegaE=0, omegaV=0, TimeS=0){
  KalmanMat <- matrix(c(0,0,0,0), ncol=4, nrow=101)
  for(i in 1:TimeS){
    # Prediction update
    KalmanMat[i+1,3] <- A %*%  KalmanMat[i,1] + B%*%u #muBar_t 
    KalmanMat[i+1,4] <- A%*%KalmanMat[i,2]%*%t(A) + omegaV #sigmaBar_t
    # Measurement update
    K_t <- KalmanMat[i+1,4] %*%t(C) %*% solve(C%*%KalmanMat[i+1,4] %*%t(C) + omegaE)
    KalmanMat[i+1,1] <- KalmanMat[i+1,3] + K_t %*% (yt[i] - C%*%KalmanMat[i+1,3]) #mu_t
    KalmanMat[i+1,2] <- (diag(ncol(K_t))-K_t%*%C) %*% KalmanMat[i+1,4]  # sigma_t
  }
  KalmanFrame <- data.frame(KalmanMat[-1,], Time=1:TimeS)
  return(KalmanFrame)
}

BS <- function(A, Kalman, x){
  x_t <- data.frame(Val=0)
  for(k in (nrow(Kalman)-1):1){
    ht <- Kalman[k, 1] + Kalman[k, 2]%*%t(A)%*%solve(Kalman[k+1, 4]) %*% (x[k+1] - Kalman[k+1, 3])
    Ht <- Kalman[k,2]-Kalman[k,2]%*%t(A)%*%solve(Kalman[k+1,4])%*%A%*%Kalman[k,2]
    x_t[k,1] <- rnorm(1, mean=ht, sd=sqrt(Ht))
  }
  return(x_t)
}

bayesReg <- function(beta0, Xt, y, omega0, cov0){
  betaHat <- solve(t(Xt)%*%Xt)%*% t(Xt) %*% as.matrix(y)
  sigma0 <- diag(omega0/cov0,2)
  betaNew <- solve(t(Xt)%*%Xt + sigma0) %*% (t(Xt)%*%Xt%*%betaHat + sigma0%*%beta0)
  sigmaNew <- t(Xt)%*%Xt + sigma0
  Vari <- omega0 * solve(sigmaNew)
  coef <- rmvnorm(n=length(y), mean=betaNew, sigma=sqrt(diag(sqrt(diag(Vari)),length(betaNew))))
  return(coef)
}

# Gibbs sampling
y = resA[101:200, 1]
x = resA[1:100, 1]
set.seed(311015)
theta <- data.frame(matrix(ncol=2, nrow=1000)) #nrow sets number of sims
theta[1,2] <- 1  #start value for theta
Xt= matrix(data = c(rep(1,100),resA[1:100,1]), ncol=2)
for(j in 1:nrow(theta)){
  KalmanRes <- KalmanF_2(yt = y, C = matrix(theta[j,2]), omegaE = 0.5, omegaV = 0.1,
                         TimeS = 100)
  NewX <- (BS(A = matrix(1), Kalman = KalmanRes, x=x))
  Xt[1:99,2] <- as.numeric(NewX[,1])
  Xt[100,2] <- KalmanRes[100,1]
  coefs <- bayesReg(beta0 = matrix(data=c(0,0), ncol=1),Xt=Xt, y=y, omega0=0.5,cov0=100)
  theta[j+1,] <- colMeans(coefs) 
}
ggplot(theta, aes(y=X2, x=1:nrow(theta))) + geom_line() + theme_classic() + labs(caption="Figure 5", title="Trace plot", x="1:1000")
ggplot(theta, aes(x=X2)) + geom_density() +  theme_classic() +
  geom_vline(xintercept = c(mean(theta$X2),0.93), col=c("red","blue")) + labs(caption="Figure 6", title=expression(Posterior~distribution~of~theta)) 
```

In the density plot is the probability for different theta values plotted. The mean of our theta equals to # Assignment 1

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dlm)
library(gridExtra)
library(mvtnorm)
```


## a) - Simulate the model

$y_t = Cx_t + \epsilon, \qquad \qquad \qquad \:\:\:\:\epsilon_t \sim  \mathcal{N}(0, \Omega_\epsilon)$  
$x_t = Ax_{t-1} + Bu_t + v_t, \qquad v_t \sim  \mathcal{N}(0, \Omega_v) \qquad x_o \sim  \mathcal{N}(\mu_0, \Sigma_0)$

Set $x_0=0$ and simulate T=100 observations from the following model. 

$y_t = 0.93x_t + \epsilon, \qquad \epsilon \sim  \mathcal{N}(0, \Omega_\epsilon=0.5)$  
$x_t = x_{t-1} + v_t, \qquad v_t \sim  \mathcal{N}(0, \Omega_v=0.1) \quad x_o \sim  \mathcal{N}(\mu_0=0, \Sigma_0=10)$   

A linear Guassian state space model is implemented in line with the given instructions. 100 observations are simulated from this model and the obtained states(x) and observations (y) are plotted in figure 1. 

```{r, echo=TRUE, fig.height=3, fig.width=7}
# Assignment 1

# a) - Simulate the model
ssm1 <- function(C=matrix(0), A=matrix(1), B=matrix(0), u=matrix(0), x0=0,
                 omegaE=0, omegaV=0, TimeS=0) {
  xt <- 0
  yt <- 0
  vt <- rnorm(1, 0, omegaV)
  xt[1] <- A%*%x0 + vt
  et <- rnorm(1, 0, omegaE)
  yt[1] <- C%*%xt + B%*%u + et
  for(i in 2:TimeS){
    vt <- rnorm(1, 0, omegaV)
    xt[i] <- A%*%xt[i-1] + vt
    et <- rnorm(1, 0, omegaE)
    yt[i] <- C%*%xt[i]+ B%*%u + et
  }
  res <- data.frame(Val=c(xt,yt), t= rep(1:TimeS,2), 
                    Value=rep(c("State","Observation"), each=TimeS))
  return(res)
}
set.seed(1234)
resA <- ssm1(C = matrix(0.93), x0 = 0, omegaE = sqrt(0.5), omegaV = sqrt(0.1), TimeS = 100)
plotA <- ggplot(resA, aes(x=t, y=Val, col=Value)) + geom_line(size=1) + theme_classic() + 
  ylim(-4,4) + scale_color_manual(values=c("skyblue", "royalblue")) 
plotA + labs(caption = "Figure 1")
```


## b) - Filtering: Sequential state inference

The point estimates for the Kalman filter and the 0.95 probability bands for the point estimates are plotted together with the states and observations simulated in *a)*.

```{r, echo=TRUE, fig.height=3, fig.width=7, message=FALSE, warning=FALSE}
KalmanF <- function(yt, C=matrix(0), A=matrix(1),
                    B=matrix(0), u=matrix(0), omegaE=0, omegaV=0, TimeS=0){
  KalmanMat <- matrix(c(0,0), ncol=2, nrow=101)
  for(i in 1:TimeS){
    # Prediction update
    muBar_t <- A %*%  KalmanMat[i,1] + B%*%u
    sigmaBar_t <- A%*%KalmanMat[i,2]%*%t(A) + omegaV
    # Measurement update
    K_t <- sigmaBar_t%*%t(C) %*% solve(C%*%sigmaBar_t%*%t(C) + omegaE)
    KalmanMat[i+1,1] <- muBar_t + K_t %*% (yt[i] - C%*%muBar_t)
    KalmanMat[i+1,2] <- (diag(ncol(K_t))-K_t%*%C) %*% sigmaBar_t
  }
  KalmanFrame <- data.frame(KalmanMat[-1, ], Time=1:TimeS)
  return(KalmanFrame)
}

resB <- KalmanF(yt = resA[101:200,1], C = matrix(0.93), omegaE = 0.5, omegaV = 0.1, TimeS=100)
resB_1 <- data.frame(Val=c(resB[,1],resB[,1]+1.96*sqrt(resB[,2]),resB[,1]-1.96*sqrt(resB[,2])), Time=rep(1:100, 3), Value=rep(c("Kalman mean", "Upper", "Lower"), each=100))
                     
plotA + geom_line(data=resB_1, aes(x=Time, y=Val, col=Value), size=1) +ylim(-5,4)+
  scale_colour_manual(values=c("red3", "darkorange","skyblue","royalblue","darkorange"))+
  labs(x="Time", title="Kalman filter estimates with 0.95 probability intervals", caption="Figure 2")
```

In figure 3 are the Kalman filter estimates and probability intervals obtained by the implemented Kalman algorithm compared to the estimates returned by the *dlmFilter* function from the *dlm* package.   

```{r, echo=TRUE, fig.height=3, fig.width=7}
filteringB <- dlmFilter(y=resA[101:200,1], dlm(m0=0, V=0.5, W=0.1, C0=10, FF=0.93, GG=1) )

filterB <- data.frame(y=(filteringB$m)[-1])
filterVar <- unlist(dlmSvd2var(u = filteringB$U.C, d = filteringB$D.C))
filterB[101:300,1] <- c(filterB$y + 1.96*sqrt(filterVar[-1]), 
                        filterB$y - 1.96*sqrt(filterVar[-1]))
filterB$Time <- rep(1:100, 3)
filterB$Value <- rep(c("dlm Kalman mean", "dlm Upper", "dlm Lower"), each=100)

ggplot(resB_1, aes(x=Time, y=Val, col=Value,linetype=Value)) + geom_line(size=1) + theme_classic()+ ylim(-5,4)+ geom_line(data=filterB, aes(x=Time, y=y, col=Value, linetype=Value),size=1) +scale_colour_manual(values=c("purple","seagreen","seagreen","red3","darkorange","darkorange"))+scale_linetype_manual(values=c("dashed","dashed","dashed","solid","solid","solid"))+labs(title="Comparsion between dlmFilter and KalmanF function", caption="Figure 3")
```

Except for the first values are the point estimates and the probability intervals identical. 

## c) - Prediction of state and data by simulation

When simulating we use *k* equals 5 and *i* goes from 1:1000. 
$$x_{T+k}^{(i)} \sim p(x_{T+k} \mid Y_{1:T})$$
$$y_{T+k}^{(i)} \sim p(y_{T+k} \mid x_{T+k}^{(i)})$$

The probability intervals for 5 future states and observations are plotted in figure 2 together with the data plotted in figure 2. The 5 future time steps are simulated 10 000 times and a 0.95 probabilty interval is given by the quantiles from the sample. 

```{r, echo=TRUE, fig.height=3, fig.width=7}
PredFunc <- function(C=matrix(0), A=matrix(1), B=matrix(0), u=matrix(0), x0=0,
                     omegaE=0, omegaV=0, TimeS=0){
  xt <- 0
  yt <- 0
  vt <- rnorm(1, 0, omegaV)
  xt[1] <- x0 + vt
  et <- rnorm(1, 0, omegaE)
  yt[1] <- C%*%xt + B%*%u + et
  for(i in 2:TimeS){
    vt <- rnorm(1, 0, omegaV)
    xt[i] <- A%*%xt[i-1] + vt
    et <- rnorm(1, 0, omegaE)
    yt[i] <- C%*%xt[i]+ B%*%u + et
  }
  res <- data.frame(x=c(xt, y=yt))
  return(res)
}

PredC <- as.data.frame(matrix(nrow=10,ncol=10000))
for(i in 1:10000){
  PredC[,i] <- PredFunc(C = matrix(0.93), x0 = rnorm(1, mean=resB$X1[100], 
                        sd=sqrt(resB$X2[100])),omegaE = sqrt(0.5), omegaV = sqrt(0.1),
                        TimeS = 5)
}
quanTs <- t(apply(PredC, 1, quantile, probs=c(0.025, 0.975)))
quanTsFrame <- data.frame(x=c(quanTs[,1], quanTs[,2]), Time= 100+rep(1:5, 4),
                 Var=rep(c("x", "y"), 2,each=5), InterV=rep(c("Lower", "Upper"), each=10))
quanTsFrame$Interval <- interaction(quanTsFrame$Var, quanTsFrame$InterV)
quanTsFrame$Interval<- factor(quanTsFrame$Interval, levels=rev(levels(quanTsFrame$Interval)))

ggplot(resA, aes(x=t, y=Val, col=Value)) + geom_line(size=1) + theme_classic() + ylim(-5,4) +
  geom_line(data=resB_1[1:100,], aes(x=Time, y=Val), col="seagreen", size=1.1) +
  geom_line(data=resB_1[101:200,], aes(x=Time, y=Val), col="darkorange", size=1) +
  geom_line(data=resB_1[201:300,], aes(x=Time, y=Val), col="darkorange", size=1) +
  geom_line(data=quanTsFrame, aes(x=Time, y=x, col=Interval), size=1, linetype="dashed") +
  scale_color_manual(values=c("skyblue", "royalblue","darkmagenta","darkmagenta","red3", "red3"))+ labs(x="Time", title="0.95 probability intervals for k=5", caption="Figure 4")
```

We notice how the width of these bands increase with time. This implies that we become more unsure for more distant future time steps.

## d) - State and model parameter inference

Given the information from the lab instructions we create our FFBS function in order to be able to update the x values.
We implement a bayesian linear regression to be able to update our $\theta$ values. This will be used in our gibbs function and the update process will continue iteratively for 1000 iterations. These iterations can be seen below. It seems to converged rather quickly.

```{r, echo=TRUE, fig.height=3, fig.width=7}
KalmanF_2 <- function(yt, C=matrix(0), A=matrix(1),
                    B=matrix(0), u=matrix(0), omegaE=0, omegaV=0, TimeS=0){
  KalmanMat <- matrix(c(0,0,0,0), ncol=4, nrow=101)
  for(i in 1:TimeS){
    # Prediction update
    KalmanMat[i+1,3] <- A %*%  KalmanMat[i,1] + B%*%u #muBar_t 
    KalmanMat[i+1,4] <- A%*%KalmanMat[i,2]%*%t(A) + omegaV #sigmaBar_t
    # Measurement update
    K_t <- KalmanMat[i+1,4] %*%t(C) %*% solve(C%*%KalmanMat[i+1,4] %*%t(C) + omegaE)
    KalmanMat[i+1,1] <- KalmanMat[i+1,3] + K_t %*% (yt[i] - C%*%KalmanMat[i+1,3]) #mu_t
    KalmanMat[i+1,2] <- (diag(ncol(K_t))-K_t%*%C) %*% KalmanMat[i+1,4]  # sigma_t
  }
  KalmanFrame <- data.frame(KalmanMat[-1,], Time=1:TimeS)
  return(KalmanFrame)
}

BS <- function(A, Kalman, x){
  x_t <- data.frame(Val=0)
  for(k in (nrow(Kalman)-1):1){
    ht <- Kalman[k, 1] + Kalman[k, 2]%*%t(A)%*%solve(Kalman[k+1, 4]) %*% (x[k+1] - Kalman[k+1, 3])
    Ht <- Kalman[k,2]-Kalman[k,2]%*%t(A)%*%solve(Kalman[k+1,4])%*%A%*%Kalman[k,2]
    x_t[k,1] <- rnorm(1, mean=ht, sd=sqrt(Ht))
  }
  return(x_t)
}

bayesReg <- function(beta0, Xt, y, omega0, cov0){
  betaHat <- solve(t(Xt)%*%Xt)%*% t(Xt) %*% as.matrix(y)
  sigma0 <- diag(omega0/cov0,2)
  betaNew <- solve(t(Xt)%*%Xt + sigma0) %*% (t(Xt)%*%Xt%*%betaHat + sigma0%*%beta0)
  sigmaNew <- t(Xt)%*%Xt + sigma0
  Vari <- omega0 * solve(sigmaNew)
  coef <- rmvnorm(n=length(y), mean=betaNew, sigma=sqrt(diag(sqrt(diag(Vari)),length(betaNew))))
  return(coef)
}

# Gibbs sampling
y = resA[101:200, 1]
x = resA[1:100, 1]
set.seed(311015)
theta <- data.frame(matrix(ncol=2, nrow=1000)) #nrow sets number of sims
theta[1,2] <- 1  #start value for theta
Xt= matrix(data = c(rep(1,100),resA[1:100,1]), ncol=2)
for(j in 1:nrow(theta)){
  KalmanRes <- KalmanF_2(yt = y, C = matrix(theta[j,2]), omegaE = 0.5, omegaV = 0.1,
                         TimeS = 100)
  NewX <- (BS(A = matrix(1), Kalman = KalmanRes, x=x))
  Xt[1:99,2] <- as.numeric(NewX[,1])
  Xt[100,2] <- KalmanRes[100,1]
  coefs <- bayesReg(beta0 = matrix(data=c(0,0), ncol=1),Xt=Xt, y=y, omega0=0.5,cov0=100)
  theta[j+1,] <- colMeans(coefs) 
}
ggplot(theta, aes(y=X2, x=1:nrow(theta))) + geom_line() + theme_classic() + labs(caption="Figure 5", title="Trace plot", x="1:1000")
ggplot(theta, aes(x=X2)) + geom_density() +  theme_classic() +
  geom_vline(xintercept = c(mean(theta$X2),0.93), col=c("red","blue")) + labs(caption="Figure 6", title=expression(Posterior~distribution~of~theta)) 
```

In the density plot is the probability for different theta values plotted. The mean of our theta equals to $`r mean(theta[,2])`$ which is close to our true value. The true value, 0.93, could be obtained but the probability for this to happen is low.


# Assignment 2


## a) - Estimating the variance components by MLE.

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Assignment 2
load("C:\\Users\\Gustav\\Documents\\AdvML\\Lab4\\CAPM.Rda")

# a) - Estimating the variance components by MLE

CAPMv <- CAPM_data[,c(10,17,18)]
Zt <- matrix(CAPM_data$MARKET - CAPM_data$RKFREE)
Yt <- matrix(CAPMv$IBM - CAPMv$RKFREE)

buildLocalTrend <- function(x,data){
  V = exp(x[1])
  W = diag(exp(x[2:3]),2,2)
  return(dlm(
    m0 = c(0,1),
    C0 = diag(c(100,0.5),2),
    FF = matrix(c(1,1),1,2),     
    GG = diag(2),
    JFF = matrix(c(0,1),1,2),
    V = V,
    W = W,
    X=data))
}

initVal <- c(1,1,1) # Initial values for optim on the estimated parameters 
MLEs <- dlmMLE(Yt, parm = initVal, build = buildLocalTrend, data = Zt)
dlmWithMLEs <- buildLocalTrend(MLEs$par, data = Zt)

```

By MLE the following estimates are given for the variance components. 

$\sigma_\epsilon^2$ = $`r (as.numeric(dlmWithMLEs[4]))`$,

$\sigma_{v^{(1)}}^2$ = $`r (data.frame((dlmWithMLEs[6]))[1,1])`$,

$\sigma_{v^{(2)}}^2$ = $`r (data.frame((dlmWithMLEs[6]))[2,2])`$



## b) - Filtering and smoothing

The filtering and smoothing distribution is computed for $\alpha$ and $\beta$. In the figures below are these distributions plotted together with 0.95 probability intervals. 

The respective distributions for the alpha parameter are shown in the figures below. As expected, the same pattern is seen in both figures and a more smooth curve is returned for the smoothig distribution. 

```{r, echo=TRUE, fig.height=3, fig.width=7}
## Filtering ##
Filter_2b <- dlmFilter(y = Yt, dlmWithMLEs)

# Variances
u1 <- matrix((t(data.frame(Filter_2b$U.C)))[c(seq(1,242,2)),1])
u2 <- matrix((t(data.frame(Filter_2b$U.C)))[c(seq(2,242,2)),2])
u_1 <-list()
u_2 <- list()
for(i in 1:120){
  u_1[i]<-c(u1[i+1])
  u_2[i]<-c(u2[i+1])
}
Filter_y <- data.frame(alpha=c(Filter_2b$m)[2:121],beta=c(Filter_2b$m)[123:242])
Filter_y$alphaVar <- unlist(dlmSvd2var(u = u_1, d = matrix(Filter_2b$D.C[-1,1])))
Filter_y$betaVar <- unlist(dlmSvd2var(u = u_2, d = matrix(Filter_2b$D.C[-1,2])))
# Intervals
Filter_y$alphaLower <- Filter_y$alpha - 1.96*sqrt(Filter_y$alphaVar)
Filter_y$alphaUpper <- Filter_y$alpha + 1.96*sqrt(Filter_y$alphaVar)
Filter_y$betaLower <- Filter_y$beta - 1.96*sqrt(Filter_y$betaVar)
Filter_y$betaUpper <- Filter_y$beta + 1.96*sqrt(Filter_y$betaVar)

alphaFilter <- ggplot(Filter_y, aes(x=1:120, y=alpha)) + geom_line(col="seagreen", size=1) + theme_classic() + labs(title="Alpha filtered", caption="Figure 7") +
  geom_line(data=Filter_y, aes(x=1:120, y=alphaLower), col="darkorange", size=1) +
  geom_line(data=Filter_y, aes(x=1:120, y=alphaUpper), col="darkorange", size=1) 

betaFilter <- ggplot(Filter_y, aes(x=1:120, y=beta)) + geom_line( col="seagreen", size=1) + theme_classic()+ labs(title="Beta filtered", caption="Figure 9") +
  geom_line(data=Filter_y, aes(x=1:120, y=betaLower), col="darkorange", size=1) +
  geom_line(data=Filter_y, aes(x=1:120, y=betaUpper), col="darkorange", size=1)

#### Smoothed ####
Smooth_2b <- dlmSmooth(y = Yt, dlmWithMLEs)
Smooth_frame <- data.frame(alpha=Smooth_2b$s[-1,1], beta=Smooth_2b$s[-1,2])

covList = dlmSvd2var(Smooth_2b$U.S, Smooth_2b$D.S)
varMat = t(sapply(covList, FUN=function(x) sqrt(diag(x))))

Smooth_frame$alphaSd <- varMat[-1,1]
Smooth_frame$betaSd <- varMat[-1,2]
# Intervals
Smooth_frame$alphaLower <- Smooth_frame$alpha - 1.96*Smooth_frame$alphaSd
Smooth_frame$alphaUpper <- Smooth_frame$alpha + 1.96*Smooth_frame$alphaSd
Smooth_frame$betaLower <- Smooth_frame$beta - 1.96*Smooth_frame$betaSd
Smooth_frame$betaUpper <- Smooth_frame$beta + 1.96*Smooth_frame$betaSd

alphaSmooth <- ggplot(Smooth_frame, aes(x=1:120, y=alpha)) + geom_line(size=1, col="seagreen") + theme_classic() +labs(title="Alpha smoothed", caption="Figure 8") +
  geom_line(data=Smooth_frame, aes(x=1:120, y=alphaLower), size=1, col="darkorange") +
  geom_line(data=Smooth_frame, aes(x=1:120, y=alphaUpper), size=1, col="darkorange") 

betaSmooth <- ggplot(Smooth_frame, aes(x=1:120, y=beta)) + geom_line(size=1, col="seagreen")+theme_classic()+labs(title="Beta smoothed", caption="Figure 10") +
  geom_line(data=Smooth_frame, aes(x=1:120, y=betaLower),size=1 ,col="darkorange") +
  geom_line(data=Smooth_frame, aes(x=1:120, y=betaUpper),size=1, col="darkorange") 

grid.arrange(alphaFilter, alphaSmooth, ncol=2)
```

For the beta parameter is the pattern a little bit different for the respective distribution. In figure 10, for the smooth distribution, the value is increasing from around time point 50 and onwards. This pattern was not seen in the figure for the filter distribution. 

```{r, echo=FALSE, fig.height=3, fig.width=7}
grid.arrange(betaFilter, betaSmooth, ncol=2)
```


## c) - Average sensitivity pre and post 1982 - a retrospective analysis

The retrospective analysis is performed by first computing the filtering distribution for the time points pre 1982. This is done by only using data for the first 48 observed time steps. Then, the same thing is done once again but for the data points from time point 49 to 120.

The *dlmBSample* function is used for making the *BS* step in the *FFBS* algorithm on the respective filtering distribution. With this approach are the mean values presented below obtained for the respective time period. Hence, it is concluded that the average sensitivity was higher after 82 than before 82. 

```{r, echo=TRUE, fig.height=3, fig.width=7}
filter_1 <- dlmFilter(Yt[1:48], buildLocalTrend(MLEs$par, Zt[1:48]))
filter_2 <- dlmFilter(Yt[49:120], buildLocalTrend(MLEs$par, Zt[49:120]))

set.seed(311015)
mean_1 <- mean(dlmBSample(filter_1)[,2])
mean_2 <- mean(dlmBSample(filter_2)[,2])

cat("Mean before 82:", mean_1)
cat("Mean after 82:", mean_2)
```


## d) - Bayesian inference for the variance components


The first way is to use a likelihood that integrates out the states (i.e. will only be a function of the three sigma you have). This likelihood is then multiplied by a prior for the unknown's sigmas. This posterior can be sampled from using, for example, Metropolis-Hastings.

 See slide 6, Lecture 2 of how a likelihood a LGSS looks when it integrated the states. 



# Appendix - R-code
```{r code=readLines(knitr::purl('C:/Users/Gustav/Documents/AdvML/Lab4/Lab_4.rmd', documentation = 0)), eval = FALSE, tidy=TRUE}
```


